# Comparing Two Population Means: Independent Samples

## Overview

Lecture 6 considered situations where observations from two conditions were **meaningfully linked**.

For paired data, each pair produced one difference:

$$
d_i=X_{1i}-X_{2i}
$$

and the analysis became a one-sample inference problem involving:

$$
\mu_d
$$

Now consider a different situation.

Suppose we want to compare:

- Spending by homeowners vs. renters
- Sales of national-brand vs. private-brand products
- Customers assigned to Website A vs. Website B
- Employees receiving one training program vs. employees receiving another

In these situations, the observations in one group do not have a meaningful one-to-one connection with observations in the other group.

These are **independent samples**.

Instead of analyzing within-pair differences, we compare the two sample means:

$$
\bar{X}_1-\bar{X}_2
$$

to learn about the difference between the two population means:

$$
\boxed{\mu_1-\mu_2}
$$

The major distinction is:

> **Paired samples compare within-pair differences. Independent samples compare two separate group means.**

---

# Why Study Design Still Matters

Statistical analysis involves more than calculating a p-value.

Suppose a retailer asks:

> Would offering more coupons increase customer spending?

There are several different questions hidden inside this problem.

### Describe

How does spending differ between customers who use coupons and customers who do not?

### Quantify Uncertainty

How precisely can we estimate the difference in population mean spending?

### Evaluate Causality

Does the difference actually result from coupons, or are customers who use coupons already different from customers who do not?

A hypothesis test addresses **sampling uncertainty**.

It does not automatically eliminate:

- Confounding
- Selection bias
- Differences in the underlying groups

Therefore:

> **Statistical significance and causal evidence are different questions.**

The study design determines what conclusions the statistical analysis can support.

---

# Paired vs. Independent Samples

Before comparing two means, determine whether the observations are **paired or independent**.

## Paired Samples

Observations are paired when each observation in one condition has a meaningful one-to-one connection with an observation in the other condition.

Examples:

- Same customer before and after treatment
- Same store before and after a redesign
- Identical twins
- Matched treatment and control stores

Each pair produces:

$$
d_i=X_{1i}-X_{2i}
$$

and inference is performed on:

$$
\mu_d
$$

---

## Independent Samples

Samples are independent when the groups contain **different observational units** and there is no meaningful one-to-one link between observations across groups.

Examples:

- 20 randomly selected men vs. 20 independently selected women
- Homeowners vs. renters
- National-brand products vs. private-brand products
- Customers assigned to two different website versions

Each observational unit contributes to only **one group**.

The parameter of interest is:

$$
\boxed{\mu_1-\mu_2}
$$

---

# The Observational Unit Determines the Design

The **observational unit** is the individual entity on which a measurement is taken.

Before choosing a statistical test, ask:

> **What is the observational unit?**

and then:

> **Does each observational unit contribute one measurement or a linked pair of measurements?**

If each unit contributes to only one group:

$$
\boxed{\text{Independent Samples}}
$$

If the same unit contributes measurements under both conditions, or units are intentionally matched:

$$
\boxed{\text{Paired Samples}}
$$

---

## Example: Taste Test

Suppose a company wants to compare mean taste ratings for Coke and Pepsi.

### Independent-Samples Design

Randomly assign participants to one of two groups:

- Group 1 tastes Coke
- Group 2 tastes Pepsi

Each participant tastes only one product.

The groups are independent.

### Paired Design

Every participant tastes both Coke and Pepsi.

Each participant provides:

$$
(\text{Coke Rating}_i,\text{Pepsi Rating}_i)
$$

The observations are paired because both ratings come from the same person.

The two designs answer similar questions but use different information.

The paired design controls for individual differences in taste preferences because each participant acts as their own comparison.

However, the study would need to prevent **order effects**.

For example, tasting Coke first might affect a participant's Pepsi rating.

A common solution would be to randomize which product each participant tastes first.

---

# Independent Samples and the Population Difference

Suppose Group 1 has population mean:

$$
\mu_1
$$

and Group 2 has population mean:

$$
\mu_2
$$

The parameter of interest is:

$$
\boxed{\Delta=\mu_1-\mu_2}
$$

The sample estimate is:

$$
\boxed{\hat{\Delta}=\bar{X}_1-\bar{X}_2}
$$

Therefore:

$$
\bar{X}_1-\bar{X}_2
\rightarrow
\text{estimate of }
\mu_1-\mu_2
$$

The direction must be defined **before interpreting the result**.

For example, suppose:

$$
\Delta
=
\mu_{\text{Homeowners}}
-
\mu_{\text{Renters}}
$$

Then:

- $\Delta>0$ means homeowners spend more on average
- $\Delta<0$ means renters spend more on average
- $\Delta=0$ means the population means are equal

Reversing the order reverses the sign:

$$
\mu_1-\mu_2
=
-(\mu_2-\mu_1)
$$

Therefore:

> **Always define the direction of the difference before interpreting it.**

---

# Hypotheses for Two Independent Means

The most common null hypothesis is:

$$
H_0:\mu_1-\mu_2=0
$$

which is equivalent to:

$$
H_0:\mu_1=\mu_2
$$

The alternative depends on the research question.

### Two-Sided

To determine whether the population means differ:

$$
H_a:\mu_1-\mu_2\neq0
$$

### Right-Tailed

To determine whether Population 1 has a larger mean:

$$
H_a:\mu_1-\mu_2>0
$$

### Left-Tailed

To determine whether Population 1 has a smaller mean:

$$
H_a:\mu_1-\mu_2<0
$$

Again, the meaning of positive and negative depends on how Group 1 and Group 2 were defined.

---

# Two Samples, Two Sources of Uncertainty

With paired samples, uncertainty comes from the variation among the within-pair differences.

With independent samples, we estimate:

$$
\mu_1
$$

using:

$$
\bar{X}_1
$$

and independently estimate:

$$
\mu_2
$$

using:

$$
\bar{X}_2
$$

Both estimates contain sampling uncertainty.

Therefore:

$$
\bar{X}_1-\bar{X}_2
$$

has uncertainty from **both samples**.

For independent samples:

$$
SE(\bar{X}_1-\bar{X}_2)
=
\sqrt{
\frac{\sigma_1^2}{n_1}
+
\frac{\sigma_2^2}{n_2}
}
$$

In practice, the population standard deviations are usually unknown, so they must be estimated from the samples.

An important consequence is:

> **The less precise group contributes more uncertainty to the comparison.**

For example, if:

$$
n_1=10{,}000
$$

but:

$$
n_2=20
$$

the second sample will generally contribute much more uncertainty to the estimated difference.

A huge sample in one group does not eliminate uncertainty created by a small sample in the other.

---

# Two-Sample t-Tests

There are two major approaches for comparing two independent population means:

1. **Welch's two-sample t-test**
2. **Pooled two-sample t-test**

Both estimate the same population quantity:

$$
\mu_1-\mu_2
$$

The difference is how they estimate the uncertainty.

---

# Welch's Two-Sample t-Test

**Welch's t-test** estimates the variance of each population separately.

It does **not** assume that the two populations have equal variances.

The estimated population difference is:

$$
\hat{\Delta}
=
\bar{X}_1-\bar{X}_2
$$

The standard error is:

$$
\boxed{
SE(\hat{\Delta})
=
\sqrt{
\frac{s_1^2}{n_1}
+
\frac{s_2^2}{n_2}
}
}
$$

where:

$s_1$ = sample standard deviation of Group 1  
$s_2$ = sample standard deviation of Group 2  
$n_1$ = sample size of Group 1  
$n_2$ = sample size of Group 2

The test statistic is:

$$
\boxed{
t=
\frac{
(\bar{X}_1-\bar{X}_2)-0
}{
\sqrt{
\frac{s_1^2}{n_1}
+
\frac{s_2^2}{n_2}
}
}
}
$$

More generally, if the null hypothesis specifies some difference:

$$
\Delta_0
$$

then:

$$
t=
\frac{
\hat{\Delta}-\Delta_0
}{
SE(\hat{\Delta})
}
$$

As before, the t-statistic represents:

> **How many estimated standard errors the observed difference lies from the null value.**

---

## Welch's Test in R

R uses Welch's method by default when performing an independent two-sample t-test.

```{r}
t.test(
    outcome ~ group,
    data = my_data
)
```

For example:

```{r}
t.test(
    spending ~ homeowner_desc,
    data = households
)
```

Unless otherwise specified, `t.test()` does **not** assume equal population variances.

This makes Welch's method the standard default when there is no strong reason to assume that the two population variances are equal.

---

# Pooled Two-Sample t-Test

The **pooled t-test** makes an additional assumption:

$$
\boxed{
\sigma_1^2=\sigma_2^2=\sigma^2
}
$$

In other words, the two populations are assumed to have the **same variance**.

Because both samples are assumed to estimate the same population variance, their information can be combined into a single **pooled variance estimate**.

The pooled variance is:

$$
\boxed{
s_p^2
=
\frac{
(n_1-1)s_1^2
+
(n_2-1)s_2^2
}{
n_1+n_2-2
}
}
$$

The pooled standard deviation is:

$$
s_p=\sqrt{s_p^2}
$$

The standard error becomes:

$$
\boxed{
SE_{\text{pooled}}
=
s_p
\sqrt{
\frac{1}{n_1}
+
\frac{1}{n_2}
}
}
$$

The degrees of freedom are:

$$
df=n_1+n_2-2
$$

---

## Pooled Test in R

Unlike Welch's method, pooling must be requested explicitly:

```{r}
t.test(
    outcome ~ group,
    data = my_data,
    var.equal = TRUE
)
```

The argument:

```{r}
var.equal = TRUE
```

tells R to assume equal population variances.

---

# Welch vs. Pooled t-Test

Both methods estimate:

$$
\mu_1-\mu_2
$$

but they make different assumptions about variability.

| Feature | Welch | Pooled |
|---|---|---|
| Compares two independent means | Yes | Yes |
| Assumes equal population variances | No | Yes |
| Estimates variances separately | Yes | No |
| Combines variance estimates | No | Yes |
| R `t.test()` default | Yes | No |

The practical rule is:

> **Use Welch's method by default unless there is a defensible reason to assume equal population variances.**

The pooled method is appropriate only when the equal-variance assumption has a reasonable basis.

---

# Why Unequal Sample Sizes Matter

Unequal sample sizes are especially important when the groups also have different variances.

Recall Welch's standard error:

$$
SE(\hat{\Delta})
=
\sqrt{
\frac{s_1^2}{n_1}
+
\frac{s_2^2}{n_2}
}
$$

Each group's contribution depends on both:

- Its variability
- Its sample size

A group with:

- High variability
- Small sample size

can contribute substantial uncertainty.

Therefore, before comparing two means, examine:

$$
n_1,\ n_2,\ s_1,\ s_2
$$

rather than looking only at the two means.

---

# Examine the Data Before Testing

Before running a two-sample t-test, summarize and visualize both groups.

Important features include:

- Sample size
- Mean
- Median
- Standard deviation
- Skewness
- Outliers or unusual observations

For example:

```{r}
my_data |>
    group_by(group) |>
    summarise(
        n = n(),
        mean = mean(outcome, na.rm = TRUE),
        median = median(outcome, na.rm = TRUE),
        sd = sd(outcome, na.rm = TRUE)
    )
```

A useful visualization is a boxplot:

```{r}
ggplot(
    my_data,
    aes(
        x = group,
        y = outcome
    )
) +
    geom_boxplot()
```

The statistical test should not replace descriptive analysis.

Instead:

$$
\text{Describe}
\rightarrow
\text{Check}
\rightarrow
\text{Infer}
$$

---

# Example: National vs. Private-Brand Soft Drinks

Suppose we compare total observed sales for:

- National-brand 2-liter soft drinks
- Private-brand 2-liter soft drinks

The observational unit is a:

> **Soft-drink product**

Each product belongs to either:

- National brand

or:

- Private brand

A national-brand product is not meaningfully paired with one particular private-brand product.

Therefore, the samples are:

$$
\boxed{\text{Independent}}
$$

---

## Descriptive Statistics

Suppose the data show:

| Brand | Products | Mean | Median | SD |
|---|---:|---:|---:|---:|
| National | 175 | 203.9 | 4.5 | 664.7 |
| Private | 57 | 170.4 | 62.1 | 263.9 |

Several features immediately deserve attention.

### Unequal Sample Sizes

$$
n_{\text{National}}=175
$$

while:

$$
n_{\text{Private}}=57
$$

### Unequal Variability

$$
s_{\text{National}}=664.7
$$

while:

$$
s_{\text{Private}}=263.9
$$

The national-brand group is much more variable.

### Strong Skewness

For national brands:

$$
\text{Mean}=203.9
$$

while:

$$
\text{Median}=4.5
$$

The enormous difference between the mean and median indicates a highly right-skewed distribution with some extremely large observations.

This is why descriptive analysis should occur **before** inferential analysis.

---

# Applying Welch's Test

Define the population difference as:

$$
\Delta
=
\mu_{\text{National}}
-
\mu_{\text{Private}}
$$

Suppose the estimated means are:

$$
\bar{X}_{\text{National}}
=
203.89
$$

and:

$$
\bar{X}_{\text{Private}}
=
170.41
$$

Then:

$$
\hat{\Delta}
=
203.89-170.41
$$

$$
\boxed{
\hat{\Delta}=33.48
}
$$

So national-brand products have an observed mean sales value approximately 33.48 units higher than private-brand products.

But the estimate alone does not tell us how uncertain that difference is.

---

# Confidence Interval for Two Independent Means

A confidence interval for the population mean difference has the general form:

$$
\boxed{
\hat{\Delta}
\pm
t^*SE(\hat{\Delta})
}
$$

For Welch's method:

$$
\boxed{
(\bar{X}_1-\bar{X}_2)
\pm
t^*
\sqrt{
\frac{s_1^2}{n_1}
+
\frac{s_2^2}{n_2}
}
}
$$

Suppose the soft-drink example produces:

$$
95\%\text{ CI}
=
(-87.14,\ 154.10)
$$

This means the data are compatible with population mean differences ranging from approximately:

$$
-87.14
$$

to:

$$
154.10
$$

under the method's assumptions.

Because the interval contains:

$$
0
$$

a population mean difference of zero remains compatible with the data.

---

# Confidence Intervals and Hypothesis Tests

For the two-sided hypotheses:

$$
H_0:\mu_1-\mu_2=0
$$

$$
H_a:\mu_1-\mu_2\neq0
$$

the confidence interval and hypothesis test communicate closely related information.

### If the 95% Confidence Interval Excludes 0

There is evidence against:

$$
H_0:\mu_1-\mu_2=0
$$

at the corresponding 5% significance level.

### If the 95% Confidence Interval Contains 0

We fail to reject:

$$
H_0:\mu_1-\mu_2=0
$$

at the corresponding 5% significance level.

For the soft-drink example:

$$
(-87.14,\ 154.10)
$$

contains:

$$
0
$$

so we fail to reject the null hypothesis.

---

# Failing to Reject Does Not Prove Equality

This distinction is essential.

Suppose:

$$
p=0.58
$$

and:

$$
95\%\text{ CI}
=
(-87.14,\ 154.10)
$$

We can say:

> The data do not provide sufficient evidence that the two population means differ.

We should **not** say:

> The population means are equal.

Why?

Because the confidence interval contains many possible differences besides zero.

For example, values such as:

$$
-50,\ 25,\ 100
$$

are also inside the interval.

The data are therefore compatible with:

- Negative differences
- No difference
- Positive differences

The analysis has not demonstrated equality.

> **Failing to find evidence of a difference is not the same as finding evidence of no difference.**

---

# Statistical Significance vs. Practical Importance

Suppose a test finds:

$$
p<0.05
$$

That tells us the observed difference would be difficult to explain purely through sampling variation under the null model.

It does **not** tell us whether the difference is important for the business.

For example, a very large sample might detect:

$$
\$0.03
$$

of additional average spending.

That difference might be statistically detectable but operationally irrelevant.

Therefore, interpretation should consider:

1. Estimated difference
2. Confidence interval
3. Statistical evidence
4. Practical magnitude
5. Study design

A p-value alone does not answer all of these questions.

---

# Statistical Evidence vs. Causal Evidence

Suppose observational data show:

$$
\bar{X}_{\text{Coupon}}
>
\bar{X}_{\text{No Coupon}}
$$

Even if a two-sample t-test finds a statistically significant difference, we cannot automatically conclude:

> Coupons caused customers to spend more.

Customers who use coupons may systematically differ from customers who do not.

For example, coupon users might:

- Shop more frequently
- Purchase larger baskets
- Be more loyal customers
- Differ demographically

A hypothesis test quantifies **sampling uncertainty**.

It does not automatically remove **confounding**.

Therefore:

> **Inference about a difference and inference about causation are separate issues.**

---

# Example: Homeowners vs. Renters

Suppose the business question is:

> How does annual Kroger spending differ between households that own their home and households that rent?

First define the populations.

Population 1:

> Kroger households that own their home

Population 2:

> Kroger households that rent their home

Define:

$$
\Delta
=
\mu_{\text{Owner}}
-
\mu_{\text{Renter}}
$$

The observational unit is:

> **Household**

Each household belongs to one group.

There is no natural one-to-one relationship between an owner household and a renter household.

Therefore:

$$
\boxed{\text{Independent Samples}}
$$

---

## Analysis Workflow in R

First summarize the groups:

```{r}
households |>
    group_by(homeowner_desc) |>
    summarise(
        n = n(),
        mean_spend = mean(spend, na.rm = TRUE),
        median_spend = median(spend, na.rm = TRUE),
        sd_spend = sd(spend, na.rm = TRUE)
    )
```

Then visualize:

```{r}
ggplot(
    households,
    aes(
        x = homeowner_desc,
        y = spend
    )
) +
    geom_boxplot()
```

Then perform Welch's two-sample t-test:

```{r}
t.test(
    spend ~ homeowner_desc,
    data = households
)
```

The output provides:

- Group means
- Estimated difference
- t-statistic
- Degrees of freedom
- p-value
- Confidence interval

The final interpretation should report the estimated population difference and its uncertainty in the context of homeowner and renter households.

---

# Paired vs. Independent t-Tests

The two procedures are closely related but solve different problems.

| Feature | Paired t-Test | Independent t-Test |
|---|---|---|
| Groups | Meaningfully linked | Separate |
| Observational units | Same/matched units | Different units |
| Parameter | $\mu_d$ | $\mu_1-\mu_2$ |
| Statistic | $\bar{d}$ | $\bar{X}_1-\bar{X}_2$ |
| Source of uncertainty | Variation in differences | Variation from both samples |
| Analysis | One-sample t-test on differences | Two-sample t-test |
| Typical R argument | `paired = TRUE` | Welch default |

The first question should therefore **not** be:

> Which t-test formula do I use?

It should be:

> **How were the data generated, and are these observations meaningfully linked?**

---

# Independent-Samples Assumptions

For a two-sample t procedure, the analysis requires the observations to have an appropriate independence structure.

Most importantly:

### Independence Between Groups

The observations in one group should not be paired with observations in the other group.

If meaningful pairing exists, use a paired analysis instead.

### Independence Within Groups

Observations should represent independent observational units rather than repeated or clustered measurements being incorrectly treated as separate independent observations.

### Distribution Shape

The behavior of the t procedure also depends on sample size and the distributions of the data.

Before inference, inspect:

- Skewness
- Extreme observations
- Sample sizes
- Variability

Large samples generally make inference about means more robust to moderate non-Normality, but extreme skewness and unusual observations should still be investigated.

### Variance Assumption

Welch's test does **not** require:

$$
\sigma_1^2=\sigma_2^2
$$

The pooled test does.

This is one reason Welch's method is generally preferred as the default.

---

# Independent-Means Analysis: Full Workflow

When comparing two population means from separate groups:

## Step 1: Identify the Observational Unit

Ask:

> What does one row or measurement represent?

Determine whether the two groups contain genuinely separate observational units.

---

## Step 2: Determine Whether the Samples Are Paired or Independent

Ask:

> Can each observation in one group be meaningfully linked to exactly one observation in the other?

If yes:

$$
\text{Paired}
$$

If no:

$$
\text{Independent}
$$

---

## Step 3: Define the Population Difference

Explicitly define:

$$
\Delta=\mu_1-\mu_2
$$

before calculating or interpreting the estimate.

This establishes what positive and negative values mean.

---

## Step 4: Examine the Data

For each group, examine:

- Sample size
- Mean
- Median
- Standard deviation
- Distribution shape
- Skewness
- Unusual observations

Use both numerical summaries and visualizations.

---

## Step 5: Estimate the Difference

Calculate:

$$
\hat{\Delta}
=
\bar{X}_1-\bar{X}_2
$$

---

## Step 6: Quantify the Uncertainty

Using Welch's method:

$$
SE(\hat{\Delta})
=
\sqrt{
\frac{s_1^2}{n_1}
+
\frac{s_2^2}{n_2}
}
$$

Construct a confidence interval:

$$
\hat{\Delta}
\pm
t^*SE(\hat{\Delta})
$$

---

## Step 7: Perform the Hypothesis Test

Usually:

$$
H_0:\mu_1-\mu_2=0
$$

against an appropriate alternative.

Use Welch's method by default:

```{r}
t.test(
    outcome ~ group,
    data = my_data
)
```

Use the pooled method only when the equal-population-variance assumption has a defensible basis:

```{r}
t.test(
    outcome ~ group,
    data = my_data,
    var.equal = TRUE
)
```

---

## Step 8: Interpret in Context

Report:

1. The direction of the comparison
2. The estimated mean difference
3. The confidence interval
4. The statistical evidence
5. Practical importance when relevant
6. Whether the study design supports a causal interpretation

The statistical result should always be translated back into the original business question.

---

# Overall Framework

The sequence from Lectures 5–7 can now be viewed as one unified framework.

### One Population Mean

Parameter:

$$
\mu
$$

Estimate:

$$
\bar{X}
$$

Method:

$$
\text{One-Sample }t
$$

### Two Linked Conditions

Parameter:

$$
\mu_d
$$

Estimate:

$$
\bar{d}
$$

Method:

$$
\text{Paired }t
=
\text{One-Sample }t\text{ on Differences}
$$

### Two Independent Populations

Parameter:

$$
\mu_1-\mu_2
$$

Estimate:

$$
\bar{X}_1-\bar{X}_2
$$

Method:

$$
\text{Welch Two-Sample }t
$$

The decision process is therefore:

$$
\boxed{
\begin{array}{c}
\text{What population quantity are we studying?}\\[6pt]
\downarrow\\
\text{One Mean}
\rightarrow
\mu
\rightarrow
\text{One-Sample }t\\[8pt]
\text{Two Means}
\rightarrow
\begin{cases}
\text{Linked observations}
\rightarrow
\mu_d
\rightarrow
\text{Paired }t\\[6pt]
\text{Separate observations}
\rightarrow
\mu_1-\mu_2
\rightarrow
\text{Welch Two-Sample }t
\end{cases}
\end{array}
}
$$

The key principle is:

> **The study design determines the statistical method.**

Do not begin by choosing a formula.

Begin by identifying:

1. The business question
2. The observational unit
3. The populations being compared
4. Whether observations are linked or independent
5. The population parameter being estimated

Then choose the statistical method that matches that structure.