# Comparing Two Population Means: Paired Samples

## Overview

Lecture 5 introduced inference for **one population parameter**.

For example, we could ask whether a population mean differs from some benchmark:

$$
H_0:\mu=\mu_0
$$

Now suppose we have **two sets of measurements** and want to compare their means.

Examples include:

- Customer spending before vs. after joining a loyalty program
- Store sales before vs. after a layout change
- Measurements from matched treatment and control stores
- Measurements from twins or other naturally matched subjects

Before choosing a statistical test, the most important question is:

> **Are the observations in the two conditions independent, or are they meaningfully connected?**

If observations can be meaningfully linked one-to-one, we have **paired data**.

If no meaningful one-to-one relationship exists, we instead have **independent samples**.

This distinction determines how the data should be analyzed.

---

# Paired vs. Independent Samples

Suppose a retailer measures checkout time at 20 stores.

The company then installs a new checkout layout and measures checkout time at the **same 20 stores again**.

For every store, there are two measurements:

$$
(\text{Before}_i,\text{After}_i)
$$

These observations are **paired** because each after measurement corresponds to the same store's before measurement.

It would be inappropriate to ignore this connection and treat the 40 measurements as unrelated observations.

Instead, calculate the change within each store:

$$
d_i=\text{After}_i-\text{Before}_i
$$

Now the original problem:

$$
\text{Before vs. After}
$$

becomes:

$$
d_1,d_2,\ldots,d_n
$$

which is a **one-sample problem involving the differences**.

---

## What Creates a Pair?

A pair exists when an observation in one condition can be meaningfully linked to exactly one observation in another condition.

Common forms of pairing include:

### Repeated Measurements

The same observational unit is measured twice.

Examples:

- Same customer before and after a program
- Same employee before and after training
- Same store before and after a redesign

### Natural Pairs

Different observational units have a meaningful natural relationship.

Examples:

- Identical twins
- Siblings
- Matched geographic locations

### Matched Pairs

Researchers intentionally match similar observational units before comparing treatments.

Examples:

- Match two stores with similar historical sales
- Match customers with similar demographics
- Match patients with similar baseline characteristics

One member of each pair may then receive one treatment while the other receives another.

---

# Why Pairing Matters

Suppose we have measurements:

$$
X_i
$$

and:

$$
Y_i
$$

for pair $i$.

Instead of analyzing the two sets separately, calculate:

$$
d_i=X_i-Y_i
$$

Each pair now contributes **one difference**.

The pair, rather than each individual measurement, becomes the **unit of analysis**.

Pairing can improve precision because characteristics shared within a pair are removed before the two conditions are compared.

For example, stores may naturally differ because of:

- Store size
- Customer traffic
- Location
- Neighborhood demographics
- Historical sales

If the same store is measured before and after a change, much of this store-to-store variation is automatically controlled because each store is compared with itself.

Conceptually:

$$
\text{Observed Difference}
=
\text{Condition Effect}
+
\text{Within-Pair Noise}
$$

rather than comparing two groups containing all of the between-unit variation.

When the measurements within pairs are meaningfully related, pairing can therefore reduce noise and improve precision.

---

# Defining the Difference

The direction of the difference matters.

Suppose:

$$
d_i=\text{After}_i-\text{Before}_i
$$

Then:

- $d_i>0$ means the measurement increased
- $d_i<0$ means the measurement decreased
- $d_i=0$ means there was no change

If the outcome is checkout time, for example:

$$
d_i<0
$$

would represent an improvement because checkout time decreased.

But if we instead defined:

$$
d_i=\text{Before}_i-\text{After}_i
$$

then improvement would produce:

$$
d_i>0
$$

There is nothing inherently correct about one direction over the other.

The important thing is to:

1. Define the difference explicitly
2. Keep the direction consistent
3. Interpret the sign in context

---

# The Population Parameter

Once the differences are defined:

$$
d_1,d_2,\ldots,d_n
$$

the parameter of interest is:

$$
\boxed{\mu_d}
$$

where:

$$
\mu_d
=
\text{population mean difference}
$$

The corresponding sample statistic is:

$$
\boxed{\bar{d}}
$$

where:

$$
\bar{d}
=
\frac{\sum d_i}{n}
$$

Therefore:

$$
\bar{d}
\rightarrow
\text{estimate of }
\mu_d
$$

This is exactly the same logic used in one-sample inference:

$$
\bar{x}\rightarrow\mu
$$

except the observations are now **within-pair differences**.

---

# Hypotheses for Paired Data

If the question is whether the two conditions differ on average, the null hypothesis is:

$$
H_0:\mu_d=0
$$

A mean difference of zero means:

> **There is no average difference between the two conditions.**

The alternative depends on the research question.

### Two-Sided

To test whether the conditions differ:

$$
H_a:\mu_d\neq0
$$

### Right-Tailed

If positive differences represent the effect of interest:

$$
H_a:\mu_d>0
$$

### Left-Tailed

If negative differences represent the effect of interest:

$$
H_a:\mu_d<0
$$

The direction depends entirely on how:

$$
d_i
$$

was defined.

---

# Paired t-Test

A **paired t-test** compares the population mean difference to a hypothesized value, usually zero.

The important insight is:

> **A paired t-test is mathematically a one-sample t-test performed on the within-pair differences.**

Once we calculate:

$$
d_i=X_i-Y_i
$$

we no longer need to think of the analysis as two separate samples.

We have one sample:

$$
d_1,d_2,\ldots,d_n
$$

and ask whether its population mean differs from zero.

---

## Standard Error of the Mean Difference

Let:

$$
s_d
$$

represent the sample standard deviation of the differences.

Then:

$$
SE(\bar{d})
=
\frac{s_d}{\sqrt{n}}
$$

where:

$n$ = number of **pairs**

Notice that $n$ is not the total number of individual measurements.

If there are 15 pairs, then:

$$
n=15
$$

even though there are:

$$
30
$$

individual measurements.

---

## Paired t-Test Statistic

To test:

$$
H_0:\mu_d=0
$$

the test statistic is:

$$
\boxed{
t=
\frac{\bar{d}-0}
{s_d/\sqrt{n}}
}
$$

More generally:

$$
t=
\frac{\bar{d}-\mu_{d,0}}
{s_d/\sqrt{n}}
$$

The degrees of freedom are:

$$
df=n-1
$$

The test statistic tells us:

> **How many standard errors the observed mean difference lies from the value specified by the null hypothesis.**

A large absolute value of $t$ means the observed mean difference is far from the null value relative to the amount of sampling uncertainty.

---

# Example: Identical Twin Study

Suppose a study compares left hippocampal volume for **15 pairs of identical twins**.

Within each pair:

- One twin was diagnosed with schizophrenia
- The other twin was not

Define:

$$
d_i
=
\text{Unaffected}_i
-
\text{Affected}_i
$$

The observed differences have:

$$
\bar{d}=0.1987
$$

and:

$$
s_d=0.2383
$$

Because:

$$
n=15
$$

the standard error is:

$$
SE(\bar{d})
=
\frac{0.2383}{\sqrt{15}}
$$

$$
SE(\bar{d})
\approx0.0615
$$

Under:

$$
H_0:\mu_d=0
$$

the test statistic is:

$$
t=
\frac{0.1987-0}
{0.0615}
$$

$$
t\approx3.23
$$

with:

$$
df=15-1=14
$$

So the observed mean difference is approximately:

$$
3.23
$$

standard errors above the null value of zero.

---

## Paired t-Test in R

If the differences have already been calculated:

```{r}
#| eval: false

t.test(
    pairs$Difference,
    mu = 0
)
```

This is simply a **one-sample t-test on the differences**.

The example produces approximately:

$$
t=3.23
$$

$$
df=14
$$

$$
p=0.006
$$

and a 95% confidence interval of approximately:

$$
(0.067,\ 0.331)
$$

Because:

$$
p<0.05
$$

we reject:

$$
H_0:\mu_d=0
$$

The data provide evidence that the population mean difference is not zero.

---

## Using `paired = TRUE`

R can also calculate the differences internally.

For example:

```{r}
#| eval: false

t.test(
    pairs$Unaffected,
    pairs$Affected,
    paired = TRUE
)
```

This performs the same conceptual analysis:

$$
d_i
=
\text{Unaffected}_i
-
\text{Affected}_i
$$

followed by a one-sample t-test on those differences.

---

# Confidence Interval for a Mean Difference

The confidence interval for the population mean difference follows the same structure as the confidence interval for a single population mean:

$$
\boxed{
\bar{d}
\pm
t^*
\frac{s_d}{\sqrt{n}}
}
$$

For the twin example, the 95% confidence interval is approximately:

$$
(0.067,\ 0.331)
$$

Because the entire interval lies above zero, the data are consistent with a positive population mean difference under the chosen definition:

$$
d_i
=
\text{Unaffected}_i
-
\text{Affected}_i
$$

The confidence interval also communicates the estimated **magnitude** of the difference, rather than merely whether the difference is statistically significant.

---

# Statistical Evidence vs. Causality

A statistically significant difference does **not automatically establish causation**.

In the twin example, the analysis provides evidence that mean left hippocampal volume differs between twins with and without a schizophrenia diagnosis.

That supports a statement about **association**:

> Mean hippocampal volume differs between the two conditions.

It does not, by itself, establish:

> Schizophrenia caused the difference in hippocampal volume.

Causal conclusions depend on the **study design**, not simply on obtaining a small p-value.

This connects back to the distinction between observational and experimental studies:

> **Statistical significance does not turn an observational study into a randomized experiment.**

---

# Matched-Pairs Experimental Design

Pairing can also be intentionally incorporated into an experiment.

Suppose Kroger has candidate stores for an endcap-display experiment.

Some stores will receive:

- **Treatment:** endcap display

while others receive:

- **Control:** usual shelf placement

Instead of randomly assigning stores without considering their characteristics, Kroger could use a **matched-pairs design**.

---

## Step 1: Match Similar Units

Before treatment assignment, form pairs of similar stores using pre-experiment characteristics such as:

- Baseline sales
- Store size
- Customer traffic
- Neighborhood characteristics

The goal is to make the stores within each pair as comparable as possible before the treatment occurs.

---

## Step 2: Randomize Within Each Pair

Within each matched pair:

- Randomly assign one store to treatment
- Randomly assign the other to control

Randomization remains important.

Matching improves comparability and precision, while random assignment supports causal inference.

---

## Step 3: Calculate One Difference Per Pair

For pair $i$, calculate:

$$
d_i
=
\text{Sales}_{\text{endcap},i}
-
\text{Sales}_{\text{control},i}
$$

Then analyze:

$$
d_1,d_2,\ldots,d_n
$$

and test:

$$
H_0:\mu_d=0
$$

The key distinction is:

> **Matching improves precision. Random assignment supports causality.**

They serve different purposes.

---

# Assumptions of the Paired t-Test

Once paired data are converted into differences:

$$
(X_i,Y_i)
\rightarrow
d_i=X_i-Y_i
$$

the assumptions of the paired t-test apply primarily to the **differences**.

The main assumptions are:

1. There is a meaningful reason for the observations to be paired
2. The differences are independent across pairs
3. The differences contain no severe outliers
4. The distribution of the differences is approximately Normal when the number of pairs is small

---

## Independence Within a Pair Is Not Required

The two observations within a pair do **not** need to be independent.

In fact, the reason for using a paired analysis is usually that they are related.

For example:

$$
(\text{Before}_i,\text{After}_i)
$$

for the same store are naturally connected.

That relationship is exactly what the paired analysis preserves.

What needs to be independent is:

$$
d_1,d_2,\ldots,d_n
$$

across separate pairs.

---

## Normality Applies to the Differences

A paired t-test does **not** require both original variables to be Normally distributed.

You do not separately need:

$$
X\sim N
$$

and:

$$
Y\sim N
$$

Instead, the relevant quantity is:

$$
D=X-Y
$$

For small samples, we want the distribution of:

$$
d_1,d_2,\ldots,d_n
$$

to be reasonably consistent with a Normal distribution and free of severe outliers.

Therefore:

> **Check the differences, not the two original columns separately.**

---

# Checking the Differences

After calculating:

```{r}
#| eval: false

pairs <- pairs |>
    mutate(
        Difference = Unaffected - Affected
    )
```

we can inspect the differences directly.

### Histogram

```{r}
#| eval: false

ggplot(pairs, aes(x = Difference)) +
    geom_histogram()
```

Look for:

- Severe skewness
- Extreme outliers
- Unusual shape

### Q–Q Plot

```{r}
#| eval: false

qqnorm(pairs$Difference)
qqline(pairs$Difference)
```

For a small sample, strong departures from the line may indicate that the Normality assumption is questionable.

However, with only a small number of pairs, visual diagnostics themselves are uncertain.

For example, with:

$$
n=15
$$

there are relatively few observations available for judging the shape of the difference distribution.

---

# Pairing Does Not Automatically Mean Independence

Paired data can introduce another important issue.

Suppose two stores are observed every week for 104 weeks.

For each week:

$$
d_t
=
\text{Sales}_{8,t}
-
\text{Sales}_{208,t}
$$

It might be tempting to treat the 104 weekly differences as:

$$
d_1,d_2,\ldots,d_{104}
$$

independent observations.

But these differences occur **over time**.

Consecutive weeks may be related.

This creates a potential violation of the independence assumption.

---

# Time Structure in Paired Differences

Suppose the difference between two stores changes systematically over time.

A single overall mean:

$$
\bar{d}
$$

might conceal that pattern.

For example, the differences might begin mostly negative and become increasingly positive.

Then the statement:

> The average difference is 10.

does not fully describe what is happening.

The relationship may instead involve a **time trend**.

This connects paired inference back to time-series analysis:

> Before reducing repeated measurements to one overall mean difference, check whether the difference itself changes systematically over time.

---

# Autocorrelation

**Autocorrelation** measures whether observations are related to observations from earlier time periods.

For weekly differences:

$$
d_t
$$

we might ask whether:

$$
d_t
$$

is related to:

$$
d_{t-1}
$$

or:

$$
d_{t-2}
$$

and so on.

An **autocorrelation function (ACF)** examines these relationships at different lags.

For example:

```{r}
#| eval: false

acf(differences)
```

A bar at lag:

$$
k
$$

measures the relationship between observations separated by:

$$
k
$$

time periods.

If substantial autocorrelation exists, the observations do not provide completely independent pieces of information.

That matters because the usual paired t-test standard error:

$$
SE(\bar{d})
=
\frac{s_d}{\sqrt{n}}
$$

assumes independent differences.

If observations are strongly dependent, treating all of them as independent can make the estimated uncertainty misleading.

---

# Changing Means vs. Autocorrelation

Time-series diagnostics can reveal different problems.

A **time plot** can reveal whether the mean difference changes over time.

An **ACF** can reveal whether observations are serially dependent.

These are not the same issue.

For example:

- A time trend means the expected difference is changing
- Autocorrelation means observations at different times are related

A dataset can have:

- A changing mean with little autocorrelation
- Autocorrelation around a stable mean
- Both
- Neither

Therefore, both the **structure of the mean** and the **dependence between observations** matter.

---

# Unit of Replication

One of the most important questions in statistical inference is:

> **What is the independent unit of replication?**

Suppose we compare two study designs.

### Study A: Many Matched Store Pairs

Suppose many pairs of similar stores participate in an experiment.

Each store is observed for six weeks.

We calculate one sales summary for each store and then one difference for each matched pair.

If there are:

$$
n
$$

independent store pairs, then the sample size is:

$$
n
$$

The independent replication comes from the **different store pairs**.

---

### Study B: One Store Pair for 104 Weeks

Suppose instead that only two stores are studied, but they are observed for:

$$
104
$$

weeks.

There are 104 weekly differences, but they all come from the **same two stores**.

Those 104 weeks are not equivalent to having 104 independent store pairs.

They provide more information about those particular stores, but not necessarily more independent replication across a population of stores.

This distinction is critical:

> **More repeated observations of the same units are not the same as more independent observational units.**

---

# Pseudoreplication

Treating repeated or dependent measurements as though they were independent observations can create **pseudoreplication**.

For example:

> Two stores observed for 104 weeks

does not automatically provide the same inferential information as:

> 104 independent pairs of stores.

Repeated observations may increase information about the particular units being studied, but they do not necessarily increase the number of independent units available for population-level inference.

The effective sample size depends on the study design and dependence structure, not simply the number of rows in the dataset.

---

# Choosing Paired vs. Independent Analysis

Before comparing two means, ask:

> **Can each observation in one condition be meaningfully linked to one observation in the other condition?**

### If Yes

The observations are paired.

Calculate:

$$
d_i=X_i-Y_i
$$

and perform **one-sample inference on the differences**.

### If No

The observations are independent samples.

Compare the two population means directly using an independent-samples method.

---

## Examples

### Same Customers Before and After

Monthly spending is recorded for the same customers before and after they join a loyalty program.

**Paired**

Each customer's before measurement is linked to that same customer's after measurement.

---

### Randomized Website Experiment

One group of customers sees the current website.

A different group sees a redesigned website.

**Independent**

There is no one-to-one correspondence between customers in the two groups.

---

### Matched Stores

Treatment stores are intentionally matched with similar control stores before an experiment.

**Paired**

Each treatment store has a specific matched control store.

---

### Different Orders Across Months

Delivery times from one group of orders this month are compared with a different group of orders last month.

**Independent**

An order this month does not correspond to one specific order from last month.

---

# Paired-Samples Workflow

When analyzing two conditions, use the following process:

1. **Identify the observational unit.**
2. Determine whether observations can be meaningfully paired.
3. If paired, define the within-pair difference:

$$
d_i=X_i-Y_i
$$

4. Determine the population parameter:

$$
\mu_d
$$

5. State the hypotheses:

$$
H_0:\mu_d=0
$$

and the appropriate alternative.

6. Check the distribution and independence of the differences.
7. Calculate:

$$
\bar{d}
$$

$$
s_d
$$

and:

$$
SE(\bar{d})
=
\frac{s_d}{\sqrt{n}}
$$

8. Calculate:

$$
t=
\frac{\bar{d}}
{s_d/\sqrt{n}}
$$

9. Obtain the p-value and confidence interval.
10. Interpret the result in the context of the study.
11. Make causal conclusions only if the study design supports them.

---

# Overall Framework

The key idea is that a paired analysis preserves information about **which observations belong together**.

Instead of:

$$
\text{Mean of Group 1}
-
\text{Mean of Group 2}
$$

while ignoring the relationships between observations, paired analysis first calculates:

$$
d_i=X_i-Y_i
$$

for each pair.

The problem then becomes:

$$
\boxed{
\text{Two Paired Conditions}
\rightarrow
\text{Within-Pair Differences}
\rightarrow
\text{One-Sample Inference}
}
$$

The major ideas are:

- **Study design determines the statistical test.**
- Pair observations only when there is a meaningful one-to-one connection.
- A paired t-test is a **one-sample t-test on differences**.
- The parameter of interest is the population mean difference:

$$
\mu_d
$$

- The sample size is the number of **independent pairs**, not the total number of measurements.
- Pairing can improve precision by removing variation between matched units.
- The assumptions apply primarily to the **differences**, not the two original columns.
- Independence is required **across pairs**, not within each pair.
- Repeated observations over time may introduce trends or dependence that a basic paired t-test does not capture.
- More repeated measurements are not equivalent to more independent units.
- Matching can improve precision, while **random assignment** is what supports a causal interpretation.

The decision rule for comparing two conditions is therefore:

$$
\boxed{
\begin{array}{c}
\text{Can observations be meaningfully linked?}\\[4pt]
\downarrow\\
\text{Yes}
\rightarrow
\text{Paired}
\rightarrow
d_i=X_i-Y_i
\rightarrow
\text{One-Sample }t\text{-Test}\\[6pt]
\text{No}
\rightarrow
\text{Independent Samples}
\end{array}
}
$$