# Descriptive Analysis

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(ggplot2)
```

## Overview

Raw data can be messy and overwhelming, making it difficult to interpret and work with. Descriptive analysis helps solve this problem by summarizing and exploring data in a way that turns raw numbers into patterns and insights that can be understood accurately and easily.

There are three key questions to consider:

- What’s typical?
- How spread out is it?
- What shape does it have?

How do we answer these questions? By examining the center, spread, and shape of the data – each of which describes a different aspect of the observations and together, helps us draw accurate conclusions about what the data is telling us.

Descriptive Analysis is organized around several key ideas:

- Center — What is typical?
- Spread — How much does the data vary?
- Shape — What does the distribution look like?
- Groups — How do different categories or populations compare?
- Visualization — How can we make patterns easier to see?

The first three, center, spread, and shape, describe the distribution itself. Groups lets you compare distributions, while visualization is how you explore and communicate those characteristics. And all are equally important.

A Quick Example: Suppose two products both sell an average of 100 units per week:

- Product A: 98, 101, 102, 99, 100 → extremely consistent
- Product B: 20, 180, 40, 160, 100 → highly variable

Both products have the same average, but their weekly sales behave very differently. Knowing the average alone does not necessarily tell us what sales actually look like.

They have the same center, but very different spread.

Together, these tools help answer the key questions needed to understand a dataset before moving on to deeper analysis.

## Center

**What's a typical value?**

Center is the measure that represents the typical or central value of a dataset. It describes the location around which the data tends to lie.

There are two main measures of center: mean and median.

### Mean
The **mean** is the arithmetic average of a dataset. It is calculated by adding all observations together and dividing by the number of observations.

$\bar{x} = \frac{\sum{x_i}}{n}$

Pros:

- Uses every observation in dataset.
- Easy to calculate and interpret.

Cons:

- Sensitive to outliers.
- Can be misleading for highly skewed distributions.


### Median
**Median** is the middle value of a dataset after the observations have been ordered. It is more robust to outliers than the mean. For a odd number of observations, sort the data and take the middle value. For an even number of observations (i.e. two middle values) take the two middle values and average them.

Pros:

- Robust to outliers and extreme values.
- Often better represents the center of skewed data.

Cons:

- Does not incorportate the magnitude of every observation the same way the mean does.
- Can ignore meaningful differences in the values away from the middle.

## Spread

**How much do values vary?**

**Spread** describes how much the observations in a dataset vary or differ from one another. In other words, it tells us how dispersed the data is around its center.

### Range
The range measures the distance between the smallest and largest observations.

$\text{range} = \text{maximum} - \text{minimum}$

Pros:

- Extremely simple to calculate or understand.

Cons:

- Only considers two observations – the largest and smallest – so it is highly sensitive to outliers.

### IQR
The **IQR** measures the spread of the middle 50% of the observations. In other words, how much do the middle 50% of values span? Because it ignores the extreme ends, it's resilient to outliers.

$\text{IQR} = Q_3 - Q_1$

$Q_1$ is the 25th percentile. 
$Q_3$ is the 75h percentile.

Conceptually, IQR looks at:
$$
\underbrace{\text{lowest 25\%}}_{\phantom{x}}
\quad|\quad
\underbrace{Q_1\; \longleftrightarrow\; Q_3}_{\text{middle 50\%}}
\quad|\quad
\underbrace{\text{highest 25\%}}_{\phantom{x}}
$$

Pros:

- Robust to outliers.

Cons:

- Does not incorportate all observations.

### Variance
**Variance** measures how spread out observations are around the mean using their squared deviations from the mean. The deviations are squared so that negative and positive deviations do not cancel each other out.

To calculate:

1. Find the mean.
2. Find how far each observation is from the mean.
3. Square each difference.
4. Add the squared differences.
5. Divide by n-1.

$s^2 = \frac{\sum{(x_i - \bar{x})^2}}{n-1}$

Pros:

- Uses every observation in the dataset.
- Measures spread specifically around the mean, so it pairs naturally with the mean.

Cons:
- Expressed in squared units, meaning it is less intuitive.

### Standard Deviation
**Standard Deviation** is the square root of the variance. It allows for an interpretable scale of how much values differ from mean by putting variance back into original units.

$s = \sqrt{s^2}$

Pros:

- Easy to interpret.

Cons:

-  Its magnitude depends on the scale of the data and can be misleading when comparing two variables of different scales.
- Sensitive to outliers.


### Coefficient of Variation
The **Coefficient of Variation** measures the standard deviation relative to the mean. In other words, it tells you how large the standard deviation is compared to the mean (how big is distance based on mean).

$\text{CV} = \frac{\text{SD}}{\bar{x}}$

Pros:

- Can accurately compare spread of variables with different scales.

Cons:

- Not useful when mean is zero, because you cannot divide by zero.
- Can be sensitive to outliers.

## Shape

**What does the distribution look like?**

**Shape** describes the overall pattern or form of a distribution. It helps us understand how observations are distributed around the center, including whether the data is symmetric, skewed, or contains outliers.

### Symmetric
A **symmetric** distribution has approximately the same shape on both sides of its center. Neither side has a substantially longer tail than the other.

In a symmetric distribution, the mean and median tend to be relatively close.

### Right Skewed
A **right-skewed** distribution has a long tail extending to the right, toward larger values. 

A relatively small number of large observations stretch the distribution to the right and can pull the mean upward.

Typically:
$\text{right skewed} \rightarrow \text{mean}>\text{median}$

### Left Skewed
A **left-skewed** distribution has a long tail extending to the left, toward smaller values.

A relatively small number of small observations stretch the distribution to the left and can pull the mean downward.

Left tail $\rightarrow$ mean pulled left $\rightarrow$ Mean < Median
Right tail $\rightarrow$ mean pulled right $\rightarrow$ Mean > Median

### Outliers
An **outlier** is an observation that is unusually far from the rest of the data.

The **IQR rule** can be used to identify potential outliers.

$\text{lower bound}=Q_1-1.5(IQR)$

$\text{upper bound}=Q_3+1.5(IQR)$

If the observation is outside of these bounds it can be flagged as a potential outlier.

An outlier should not automatically be deleted. It is important to investigate its cause first. For example, an extreme observation could represent a data-entry error, a legitimate unusual observation, or a measurement/unit issue.


## Frequency Tables
A **frequency table** summarizes a **categorical** variable by showing the count and/or proportion of observations in each category. It helps identify common and rare categories and understand the overall category mix.

| Department | Count  | Percent |
|------------|-------:|--------:|
| Grocery    | 39,021 | 42.3%   |
| Drug GM    | 31,529 | 34.1%   |
| Produce    | 3,118  | 3.4%    |


## Visualizations

| Plot | Data | Main Question |
|---|---|---|
| Histogram | 1 quantitative variable | What does the distribution look like? |
| Boxplot | Quantitative, often by category | How does the distribution/spread compare? |
| Scatterplot | 2 quantitative variables | Are the variables related? |
| Bar chart | Categorical | How do category counts/proportions compare? |
| Pie chart | Categorical proportions | How is the whole divided among categories? |

### Histograms

A **histogram** shows the **distribution** of a **single quantitative variable** by grouping numerical values into intervals called bins and showing how many observations fall into each bin.

Useful for identifying:

- Shape of a distribution
- Skewness
- Modes or peaks
- Gaps

The bin width determines how wide each interval is and therefore how much detail the histogram shows

```{r}
ggplot(mtcars, aes(x = mpg)) +
    geom_histogram(binwidth = 2)
```

**Use when: You want to see what the distribution of one quantitative variable looks like.**

### Boxplots

A **boxplot** provides a compact summary of the **distribution** of a **quantitative** variable using the five-number **summary**: $\text{Min},\ Q_1,\ \text{Median},\ Q_3,\ \text{Max}$

It is useful for examining:

- Center
- Spread
- Skewness
- Potential outliers
- Differences between groups

Boxplots are especially useful when you want to **compare distributions** across categories.

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
    geom_boxplot()
```

**Use when: You want to examine or compare the distributions of quantitative data, particularly across groups.**

### Scatterplots

A **scatterplot** shows the **relationship** between **two quantitative variables**. Each point represents one observation.

Useful for identifying:

- Direction — positive or negative relationship
- Strength — how closely the variables are related
- Form — linear, nonlinear, etc.
- Outliers

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point()
```

**Use when: You want to know whether two quantitative variables are related.**

### Bar Charts

A **bar chart** displays the counts or proportions of categories for a categorical variable. It is essentially the graphical version of a **frequency table**.

The categories are represented by bars, and the height of each bar represents its count.

```{r}
ggplot(mtcars, aes(x = factor(cyl))) +
    geom_bar()
```

**Use when: You want to compare counts or proportions across categories.**

### Pie Charts

A **pie chart** displays **proportions of a whole** as slices of a circle. Larger slices represent larger proportions.

Pie charts work best when there are only a **small number of categories**. Similar-sized slices can be difficult to compare visually, so a bar chart is often easier to interpret.

```{r}
mtcars |>
    count(cyl) |>
    ggplot(aes(x = "", y = n, fill = factor(cyl))) +
    geom_col(width = 1) +
    coord_polar(theta = "y")
```

**Use when: You want to show how a small number of categories make up a whole.**

## ggplot2
ggplot2 is a data visualization package in R built by combining data, aesthetics, and layers.

**data + aesthetics + layers**

Data = what you’re plotting
`aes()` = which variables control visual properties
`geom_*()` = how the observations are drawn
`scale_*()` = how values are displayed
`facet_*()` = split into groups
`labs()` = titles and labels
`theme_*()` = appearance

### Aesthetics
Aesthetics, specified with `aes()`, determine how variables in the data **are mapped to visual properties** of the graph.

Common aesthetics include:

- x — horizontal position
- y — vertical position
- color — color
- size — size
- shape — shape
- alpha — transparency

```{r}
ggplot(mtcars, aes(
    x = wt,
    y = mpg,
    color = factor(cyl)
))
```

### Layers

**Layers** determine how the data is actually displayed. They are usually added with `geom_*()` functions using `+`.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point()
```

Examples include:

`geom_point()`
`geom_histogram()`
`geom_boxplot()`
`geom_col()`

### Mapping vs. Setting

**Mapping** means a visual property *varies* according to a variable in the data. Put it inside `aes()`

```{r}
ggplot(mtcars, aes(
    x = wt,
    y = mpg,
    color = factor(cyl)
)) +
    geom_point()
```

Different household-size categories receive different colors.

**Setting** means giving a visual property one fixed value. Put it outside `aes()`

```{r}
ggplot(mtcars, aes(
    x = wt,
    y = mpg
    )) +
    geom_point(color = "steelblue")
```

Now every point is steel blue.

### Scales

**Scales** control how mapped data values are displayed, such as transforming or formatting an axis.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point() +
    scale_y_log10()
```

`scale_y_log10()` cisplays the y-axis on a logarithmic scale.

### Facets

**Facets** split one visualization into separate plots for different groups.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point() +
    facet_wrap(~ cyl)
```

Instead of showing every household-size group on one graph, you get a separate panel for each household size.

### Labels

`labs()` controls the descriptive text on the visualization:

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
    geom_point() +
    labs(
        title = "Vehicle Weight vs. Fuel Efficiency",
        subtitle = "1973–74 Motor Trend road tests",
        x = "Weight (1,000 lbs)",
        y = "Miles per Gallon",
        color = "Cylinders"
    )
```

### Themes

Themes control the overall appearance of the plot rather than the underlying data.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point() +
    theme_minimal()
```