# Inference for Categorical Data

## Overview

Previous inference methods focused primarily on **quantitative outcomes**.

For example:

- One population mean → $\mu$
- Paired mean difference → $\mu_d$
- Difference between two independent means → $\mu_1-\mu_2$

Lecture 8 shifts the focus to **categorical outcomes**.

Examples include:

- Purchase vs. no purchase
- Homeowner vs. renter
- Married vs. unmarried
- Store vs. website vs. app
- Current page vs. new page vs. third page

Instead of asking questions about **means**, we now ask questions about:

- **Proportions**
- **Counts**
- **Associations between categorical variables**
- **Whether observed category frequencies match an expected distribution**

The main methods are:

1. **Two-proportion inference**
2. **Chi-square test of independence**
3. **Chi-square goodness-of-fit test**

The appropriate method depends on the business question and structure of the categorical data.

---

# Two-Proportion Inference

Suppose a retailer conducts an A/B experiment.

Each visitor is randomly assigned to either:

- Current page
- New page

The outcome is whether the visitor:

- Purchases
- Does not purchase

Suppose the results are:

| Page | Visitors | Purchases | Conversion Rate |
|---|---:|---:|---:|
| Current | 10,000 | 1,200 | 12.0% |
| New | 10,000 | 1,230 | 12.3% |

The sample proportions are:

$$
\hat{p}_{Current}
=
\frac{1200}{10000}
=
0.120
$$

and:

$$
\hat{p}_{New}
=
\frac{1230}{10000}
=
0.123
$$

The estimated difference is:

$$
\hat{p}_{New}-\hat{p}_{Current}
=
0.123-0.120
=
0.003
$$

or:

$$
\boxed{0.3\text{ percentage points}}
$$

Notice that this is **0.3 percentage points**, not 0.3%.

---

## Population Parameter

The population parameter is:

$$
\boxed{
p_{New}-p_{Current}
}
$$

where:

$p_{New}$ = population conversion rate for the new page  
$p_{Current}$ = population conversion rate for the current page

The sample estimate is:

$$
\boxed{
\hat{p}_{New}-\hat{p}_{Current}
}
$$

This follows the same general inference framework used previously:

$$
\text{Sample Statistic}
\rightarrow
\text{Estimate Population Parameter}
$$

Here:

$$
\hat{p}_{New}-\hat{p}_{Current}
\rightarrow
p_{New}-p_{Current}
$$

---

# Confidence Interval for Two Proportions

A confidence interval estimates the possible values of:

$$
p_1-p_2
$$

using the observed sample proportions.

The standard error is:

$$
SE
=
\sqrt{
\frac{
\hat{p}_1(1-\hat{p}_1)
}{
n_1
}
+
\frac{
\hat{p}_2(1-\hat{p}_2)
}{
n_2
}
}
$$

A 95% confidence interval is approximately:

$$
\boxed{
(\hat{p}_1-\hat{p}_2)
\pm
1.96(SE)
}
$$

For the page experiment, the estimated difference is:

$$
0.30\text{ percentage points}
$$

and the 95% confidence interval is:

$$
\boxed{
(-0.61,\ 1.21)
\text{ percentage points}
}
$$

This interval includes:

$$
0
$$

so a population difference of zero is compatible with the data.

The interval also tells us much more than simply whether the result is statistically significant.

It describes the **range of plausible population differences**.

---

# Practical Importance

Suppose the retailer decides that implementing the new page would only be worthwhile if conversion improves by at least:

$$
0.5\text{ percentage points}
$$

The confidence interval is:

$$
(-0.61,\ 1.21)
$$

This interval includes:

- Negative effects
- No effect
- Improvements smaller than 0.5 percentage points
- Improvements larger than 0.5 percentage points

Therefore, the data do not precisely establish whether the improvement reaches the business threshold.

This illustrates an important distinction:

> **Statistical uncertainty and business importance are separate questions.**

A confidence interval allows us to compare the estimated effect with a practically meaningful threshold rather than looking only at whether:

$$
p<0.05
$$

---

# Hypothesis Test for Two Proportions

To test whether the population conversion rates differ:

$$
H_0:
p_{New}=p_{Current}
$$

which is equivalent to:

$$
H_0:
p_{New}-p_{Current}=0
$$

Against the two-sided alternative:

$$
H_a:
p_{New}\neq p_{Current}
$$

In R:

```{r}
#| eval: false
ab_test <- prop.test(
    x = c(1230, 1200),
    n = c(10000, 10000),
    alternative = "two.sided",
    correct = FALSE
)

ab_test
```

Here:

- `x` contains the number of successes
- `n` contains the total number of observations
- `alternative = "two.sided"` tests for any difference
- `correct = FALSE` disables the continuity correction

For the example:

$$
p\text{-value}=0.516
$$

Because:

$$
0.516>0.05
$$

we fail to reject:

$$
H_0
$$

The experiment does not provide sufficient evidence that the population conversion rates differ.

This does **not** prove:

$$
p_{New}=p_{Current}
$$

It means the observed data do not provide sufficient evidence of a difference.

---

# Why Proportions Are Pooled for the Hypothesis Test

There is an important difference between constructing a confidence interval and conducting the hypothesis test.

## Confidence Interval

When constructing the confidence interval, we are **not assuming the population proportions are equal**.

Therefore, we estimate their variability separately using:

$$
\hat{p}_1
$$

and:

$$
\hat{p}_2
$$

The standard error is:

$$
SE_{CI}
=
\sqrt{
\frac{\hat{p}_1(1-\hat{p}_1)}{n_1}
+
\frac{\hat{p}_2(1-\hat{p}_2)}{n_2}
}
$$

## Hypothesis Test

Under the null hypothesis:

$$
H_0:p_1=p_2
$$

we are explicitly assuming that both groups share **one common population proportion**.

Therefore, we estimate that common proportion using the pooled sample proportion:

$$
\boxed{
\hat{p}_{pool}
=
\frac{x_1+x_2}{n_1+n_2}
}
$$

The hypothesis-test standard error uses:

$$
\hat{p}_{pool}
$$

because the test asks:

> If the null hypothesis of equal population proportions were true, how much difference between the sample proportions would we expect from sampling variation alone?

So:

> **Confidence interval → keep proportions separate.**

> **Hypothesis test under $H_0:p_1=p_2$ → pool the proportions.**

---

# From Two Groups to Several Groups

Suppose the retailer introduces a third page.

The results are now:

| Page | Visitors | Purchases | Conversion Rate |
|---|---:|---:|---:|
| Current | 10,000 | 1,200 | 12.0% |
| New | 10,000 | 1,230 | 12.3% |
| Third | 15,000 | 2,100 | 14.0% |

Now the question becomes:

> **Do conversion rates differ across the three pages?**

The null hypothesis is:

$$
\boxed{
H_0:
p_{Current}
=
p_{New}
=
p_{Third}
}
$$

The alternative is:

$$
\boxed{
H_a:
\text{Not all population conversion rates are equal}
}
$$

With more than two groups, we can organize the data into a **contingency table** and use a **chi-square test**.

---

# Contingency Tables

A **contingency table**, also called a **two-way table**, displays the counts corresponding to combinations of two categorical variables.

For the page experiment, the two variables are:

1. Page
2. Purchase outcome

The table is:

| Page | Purchased | Did Not Purchase |
|---|---:|---:|
| Current | 1,200 | 8,800 |
| New | 1,230 | 8,770 |
| Third | 2,100 | 12,900 |

In R:

```{r}
#| eval: false
page_table <- cbind(
    Purchased = pages$Purchases,
    `Did not purchase` = pages$Visitors - pages$Purchases
)

rownames(page_table) <- pages$Page

page_table
```

Each visitor contributes to exactly **one cell** in the table.

---

# Why Raw Counts Alone Can Be Misleading

Notice that the third page has:

$$
2100
$$

purchases compared with:

$$
1200
$$

for the current page.

But the third page also has:

$$
15000
$$

visitors instead of:

$$
10000
$$

Therefore, comparing raw purchase counts alone would be misleading.

The relevant quantities are the **conditional proportions**:

$$
\frac{\text{Purchases}}{\text{Visitors}}
$$

which give:

- Current: 12.0%
- New: 12.3%
- Third: 14.0%

When group sizes differ, counts must be interpreted relative to the corresponding totals.

---

# Chi-Square Test of Independence

The **chi-square test of independence** determines whether two categorical variables are associated.

For the page experiment, the variables are:

- Page assignment
- Purchase outcome

The null hypothesis is:

$$
\boxed{
H_0:
\text{Page and purchase outcome are independent}
}
$$

This can equivalently be written as:

$$
\boxed{
H_0:
p_{Current}
=
p_{New}
=
p_{Third}
}
$$

The alternative is:

$$
\boxed{
H_a:
\text{Page and purchase outcome are associated}
}
$$

or equivalently:

$$
H_a:
\text{Not all population conversion rates are equal}
$$

---

# What Does Independence Mean?

Two categorical variables are **independent** if knowing the value of one does not change the distribution of the other.

In this example:

> Page and purchase outcome are independent if knowing which page a visitor sees does not change their probability of purchasing.

If:

$$
P(\text{Purchase}\mid\text{Current})
=
P(\text{Purchase}\mid\text{New})
=
P(\text{Purchase}\mid\text{Third})
$$

then page and purchase are independent.

If these probabilities differ, the variables are associated.

---

# Observed vs. Expected Counts

The chi-square test compares:

- **Observed counts** — what actually occurred
- **Expected counts** — what we would expect if the null hypothesis of independence were true

The observed counts are denoted:

$$
O_{ij}
$$

and the expected counts are denoted:

$$
E_{ij}
$$

The core question is:

> **If these variables were really independent, what would this table approximately look like?**

---

# Expected Counts Under Independence

Across all three pages:

$$
1200+1230+2100
=
4530
$$

visitors purchased.

There are:

$$
10000+10000+15000
=
35000
$$

total visitors.

Therefore, the overall purchase rate is:

$$
\frac{4530}{35000}
\approx
0.1294
$$

or:

$$
12.94\%
$$

If page and purchase outcome are independent, each page should have approximately this same conversion rate.

For the current page:

$$
E_{\text{Current,Purchased}}
=
10000(0.1294)
\approx
1294.3
$$

More generally, expected counts are calculated as:

$$
\boxed{
E_{ij}
=
\frac{
(\text{Row Total}_i)
(\text{Column Total}_j)
}{
\text{Grand Total}
}
}
$$

This formula preserves the observed row and column totals while constructing the table we would expect under independence.

---

# Why Expected Counts Can Differ Across Groups

Independence does **not** mean that every group should have identical counts.

The third page has:

$$
15000
$$

visitors while the other pages have:

$$
10000
$$

each.

Even if all three pages have exactly the same underlying conversion probability, we would expect more purchases from the third page simply because more people saw it.

Therefore:

> **Independence implies comparable proportions, not necessarily identical counts.**

---

# Pearson Chi-Square Statistic

The Pearson chi-square statistic measures how far the observed counts are from the counts expected under the null hypothesis.

$$
\boxed{
X^2
=
\sum_{i,j}
\frac{
(O_{ij}-E_{ij})^2
}{
E_{ij}
}
}
$$

Each cell contributes:

$$
\frac{(O-E)^2}{E}
$$

to the total statistic.

The logic is:

### $O-E$

Measures how far the observed count is from the expected count.

### Squaring

$$
(O-E)^2
$$

prevents positive and negative differences from canceling each other out.

### Dividing by $E$

$$
\frac{(O-E)^2}{E}
$$

scales the discrepancy relative to the size of the expected count.

Therefore:

> **Larger $X^2$ values indicate greater disagreement between the observed data and the null model.**

---

# Degrees of Freedom

For a contingency table with:

- $r$ rows
- $c$ columns

the degrees of freedom are:

$$
\boxed{
df=(r-1)(c-1)
}
$$

For the page experiment:

$$
r=3
$$

and:

$$
c=2
$$

so:

$$
df
=
(3-1)(2-1)
=
2
$$

Therefore:

$$
\boxed{df=2}
$$

Under the null hypothesis and appropriate conditions:

$$
X^2
\sim
\chi^2_{(r-1)(c-1)}
$$

approximately.

---

# Chi-Square Test in R

The test can be conducted using:

```{r}
#| eval: false
page_test <- chisq.test(
    page_table,
    correct = FALSE
)

page_test
```

For the three-page example:

$$
X^2=26.435
$$

with:

$$
df=2
$$

and:

$$
p\text{-value}
=
1.819\times10^{-6}
$$

The p-value is extremely small, so we reject:

$$
H_0
$$

and conclude that page assignment and purchase outcome are associated.

Equivalently:

> **There is evidence that at least two of the population conversion rates differ.**

---

# What an Overall Chi-Square Test Does Not Tell You

A significant chi-square test tells us:

$$
\boxed{
\text{Some association exists}
}
$$

or:

$$
\boxed{
\text{Not all population proportions are equal}
}
$$

It does **not** automatically tell us:

- Every group differs from every other group
- Which specific groups differ
- Which group should be selected by the business
- Whether the differences are practically important

For the page example, rejecting:

$$
H_0:
p_{Current}=p_{New}=p_{Third}
$$

means the equality of all three proportions is inconsistent with the data.

It does **not** mean:

$$
p_{Current}\neq p_{New}\neq p_{Third}
$$

in every pairwise comparison.

The next step is to investigate **where the association appears**.

---

# Two-Proportion Test and Chi-Square Connection

With exactly two groups and two outcomes, the contingency table is:

$$
2\times2
$$

For example:

| Page | Purchased | Did Not Purchase |
|---|---:|---:|
| New | 1230 | 8770 |
| Current | 1200 | 8800 |

The chi-square test can be run using:

```{r}
#| eval: false
two_page_table <- page_table[
    c("New", "Current"),
]

two_page_test <- chisq.test(
    two_page_table,
    correct = FALSE
)
```

For a two-sided comparison without continuity correction, the two-proportion test and chi-square test are mathematically connected:

$$
\boxed{
X^2=z^2
}
$$

Therefore, they produce the same p-value in this setting.

For this example:

$$
p\approx0.516
$$

from either method.

So the two-proportion test and chi-square test are not unrelated techniques.

The two-proportion test is closely connected to the chi-square framework when the table is:

$$
2\times2
$$

---

# Finding Where the Association Occurs

After obtaining a significant overall chi-square test, first examine the category proportions.

For the page experiment:

| Page | Conversion Rate |
|---|---:|
| Current | 12.0% |
| New | 12.3% |
| Third | 14.0% |

The pooled conversion rate is approximately:

$$
12.94\%
$$

The third page appears noticeably higher than the other two.

This provides a descriptive indication of where the association may be occurring.

However, we can investigate individual cells more formally using **standardized residuals**.

---

# Standardized Residuals

A **standardized residual** measures how far an observed cell count differs from what would be expected under independence, accounting for the cell's variability.

In R:

```{r}
#| eval: false
round(
    page_test$stdres,
    2
)
```

For the page example:

| Page | Purchased | Did Not Purchase |
|---|---:|---:|
| Current | -3.32 | 3.32 |
| New | -2.27 | 2.27 |
| Third | 5.10 | -5.10 |

Interpretation:

- **Positive residual** → more observations than expected
- **Negative residual** → fewer observations than expected
- **Large absolute residual** → stronger departure from independence

For the third page:

$$
5.10
$$

for purchases means substantially **more purchases occurred than expected under independence**.

Its non-purchase residual is:

$$
-5.10
$$

meaning substantially **fewer non-purchases occurred than expected**.

This pattern helps explain the significant overall chi-square statistic.

---

# Why Residuals Come in Opposite Directions

Within a row, the total number of visitors is fixed.

If a page has more purchases than expected, those additional observations must come from somewhere.

Therefore, it will also have fewer non-purchases than expected.

For example:

$$
O_{\text{Purchased}}
>
E_{\text{Purchased}}
$$

will correspond with:

$$
O_{\text{Not Purchased}}
<
E_{\text{Not Purchased}}
$$

when there are only two outcome categories.

---

# Conditions for the Chi-Square Approximation

Before trusting a chi-square test, check whether its assumptions are reasonable.

## Independent Observations

Observations should contribute independently.

For example, each visitor should contribute one independent observation rather than the same visitor appearing repeatedly and being treated as independent observations.

A huge dataset does not solve a violation of independence.

> **More rows do not fix dependence among observations.**

---

## Mutually Exclusive Categories

Each observational unit should contribute to exactly one cell.

For example, a visitor cannot simultaneously be classified as both:

- Purchased
- Did not purchase

The categories must be mutually exclusive for the table being analyzed.

---

## Expected Counts

Expected counts should be sufficiently large for the chi-square approximation to be reliable.

A simple conservative rule is:

$$
\boxed{
E_{ij}\geq5
}
$$

for every cell.

Expected counts can be inspected in R using:

```{r}
#| eval: false
round(
    page_test$expected,
    1
)
```

For the page example:

| Page | Purchased | Did Not Purchase |
|---|---:|---:|
| Current | 1294.3 | 8705.7 |
| New | 1294.3 | 8705.7 |
| Third | 1941.4 | 13058.6 |

All expected counts are far greater than 5.

---

# Chi-Square Test of Independence: Workflow

When determining whether two categorical variables are associated:

## Step 1: Identify the Observational Unit

Determine what one observation represents.

---

## Step 2: Identify the Two Categorical Variables

For example:

- Page
- Purchase outcome

---

## Step 3: Construct a Contingency Table

```{r}
#| eval: false
table(
    Variable1 = data$variable1,
    Variable2 = data$variable2
)
```

---

## Step 4: Examine Counts and Proportions

Do not rely only on raw counts.

Compare relevant percentages or conditional proportions.

---

## Step 5: State the Hypotheses

$$
H_0:
\text{The variables are independent}
$$

$$
H_a:
\text{The variables are associated}
$$

---

## Step 6: Calculate Expected Counts

Conceptually:

$$
E_{ij}
=
\frac{
(\text{Row Total})
(\text{Column Total})
}{
\text{Grand Total}
}
$$

---

## Step 7: Check Conditions

Verify:

- Independent observations
- Mutually exclusive categories
- Adequate expected counts

---

## Step 8: Conduct the Test

```{r}
#| eval: false
test <- chisq.test(
    contingency_table,
    correct = FALSE
)
```

---

## Step 9: Interpret the Overall Result

If the p-value is small:

> There is evidence of an association between the categorical variables.

If the p-value is not small:

> The data do not provide sufficient evidence of an association.

---

## Step 10: Investigate the Pattern

If the overall test is significant, examine:

- Conditional proportions
- Observed vs. expected counts
- Standardized residuals

```{r}
#| eval: false
test$expected
test$stdres
```

This helps determine which categories appear to contribute most strongly to the overall association.

---

# Example: Homeownership and Marital Status

Suppose we ask:

> Is homeownership associated with marital status among the households in these data?

The two categorical variables are:

### Homeownership

Possible categories include:

- Homeowner
- Probable Owner
- Probable Renter
- Renter
- Unknown

### Marital Status

The codes are:

- A = Married
- B = Unmarried
- U = Unknown

A contingency table might be created using:

```{r}
#| eval: false
table(
    Homeownership = hh$homeowner_desc,
    Marital_status = hh$marital_status_code
)
```

The hypotheses are:

$$
H_0:
\text{Homeownership and marital status are independent}
$$

$$
H_a:
\text{Homeownership and marital status are associated}
$$

Before running the test, however, we must decide which categories should be included.

---

# Category Definitions Matter

Suppose the data contain categories such as:

- Unknown homeownership
- Unknown marital status
- Probable owner
- Probable renter

The decision to include, combine, or exclude these categories changes the population represented by the analysis and therefore changes the interpretation.

For example, removing unknown marital status means the analysis is no longer about:

> All households in the dataset.

It becomes an analysis of:

> Households with known marital status.

Therefore:

> **Data-cleaning decisions are part of the statistical analysis because they determine what population the resulting inference describes.**

Categories should not be removed simply because they are inconvenient.

The decision should follow from the research question and the meaning of the categories.

---

# Association Does Not Imply Causation

Suppose the chi-square analysis finds an association between:

- Marital status
- Homeownership

Can we conclude:

> Getting married causes people to become homeowners?

No.

The chi-square test establishes statistical evidence of an **association**.

It does not establish the causal mechanism behind that association.

Other variables might affect both marital status and homeownership, such as:

- Age
- Income
- Household size
- Employment
- Geographic location

Therefore:

$$
\boxed{
\text{Association}
\neq
\text{Causation}
}
$$

The study design determines whether causal conclusions are justified.

---

# Chi-Square Goodness-of-Fit Test

The chi-square framework can also answer a different type of categorical question.

Suppose a retailer expects customers to make their first purchase through:

- Store: 50%
- Website: 30%
- App: 20%

These percentages represent a **specified benchmark distribution**.

An independent sample of 500 customers produces:

| Channel | Observed Customers | Benchmark Share |
|---|---:|---:|
| Store | 240 | 50% |
| Website | 160 | 30% |
| App | 100 | 20% |

The question is:

> **Does the observed channel mix match the retailer's specified benchmark distribution?**

This is a **chi-square goodness-of-fit test**.

---

# Independence vs. Goodness-of-Fit

These two chi-square procedures answer different questions.

## Chi-Square Test of Independence

Uses **two categorical variables**.

Example:

> Is page assignment associated with purchase outcome?

The data form a two-way contingency table.

---

## Chi-Square Goodness-of-Fit Test

Uses **one categorical variable** and compares its observed distribution with specified benchmark probabilities.

Example:

> Does customer channel mix match 50% store, 30% website, and 20% app?

The distinction is:

| Method | Variables | Question |
|---|---:|---|
| Chi-square independence | 2 categorical variables | Are the variables associated? |
| Chi-square goodness-of-fit | 1 categorical variable | Does its distribution match specified proportions? |

---

# Expected Counts for Goodness-of-Fit

Suppose:

$$
n=500
$$

and the benchmark probabilities are:

$$
p_{Store}=0.50
$$

$$
p_{Website}=0.30
$$

$$
p_{App}=0.20
$$

Expected counts are:

$$
E_i=np_{i,0}
$$

Therefore:

### Store

$$
E_{Store}
=
500(0.50)
=
250
$$

### Website

$$
E_{Website}
=
500(0.30)
=
150
$$

### App

$$
E_{App}
=
500(0.20)
=
100
$$

So under the benchmark:

| Channel | Observed | Expected |
|---|---:|---:|
| Store | 240 | 250 |
| Website | 160 | 150 |
| App | 100 | 100 |

The chi-square statistic again measures how far observed counts are from expected counts.

---

# Goodness-of-Fit Chi-Square Statistic

The statistic is:

$$
\boxed{
X^2
=
\sum_i
\frac{
(O_i-E_i)^2
}{
E_i
}
}
$$

where:

$$
E_i=np_{i,0}
$$

The logic is identical to the independence test:

> Compare what we observed with what we would expect if the null model were correct.

---

# Goodness-of-Fit Hypotheses

For the channel example:

$$
H_0:
\begin{cases}
p_{Store}=0.50\\
p_{Website}=0.30\\
p_{App}=0.20
\end{cases}
$$

The alternative is:

$$
H_a:
\text{The population distribution differs from the benchmark in at least one category}
$$

---

# Goodness-of-Fit Degrees of Freedom

If there are:

$$
k
$$

categories and their probabilities are specified in advance, then:

$$
\boxed{
df=k-1
}
$$

For three channels:

$$
df=3-1=2
$$

---

# Goodness-of-Fit in R

Suppose:

```{r}
#| eval: false
channel_counts <- c(
    Store = 240,
    Website = 160,
    App = 100
)
```

The test can be performed with:

```{r}
#| eval: false
chisq.test(
    channel_counts,
    p = c(0.50, 0.30, 0.20)
)
```

For the example:

$$
X^2=1.0667
$$

$$
df=2
$$

and:

$$
p\text{-value}=0.5866
$$

Because the p-value is not small, we fail to reject the null hypothesis.

The data do not provide sufficient evidence that the population channel distribution differs from:

$$
50\%,30\%,20\%
$$

This does **not** prove that the true population proportions are exactly:

$$
50\%,30\%,20\%
$$

It means the observed differences are not sufficiently inconsistent with the benchmark to reject it.

---

# One Logic Behind Both Chi-Square Tests

Both chi-square procedures follow the same basic structure:

$$
\boxed{
\text{Observed}
\quad\text{vs.}\quad
\text{Expected under }H_0
}
$$

and both use:

$$
\boxed{
X^2
=
\sum
\frac{(O-E)^2}{E}
}
$$

What changes is **where the expected counts come from**.

### Test of Independence

Expected counts are determined from the row and column totals:

$$
E_{ij}
=
\frac{
(\text{Row Total}_i)
(\text{Column Total}_j)
}{
\text{Grand Total}
}
$$

### Goodness-of-Fit

Expected counts come from specified benchmark probabilities:

$$
E_i=np_{i,0}
$$

So the mathematical logic is the same, but the null model is different.

---

# Choosing the Appropriate Categorical Method

The first step is identifying the business question.

| Business Question | Analysis |
|---|---|
| How much do two conversion rates differ? | Two-proportion CI and test |
| Does conversion differ across three pages? | Chi-square test on a two-way table |
| Are homeownership and marital status associated? | Chi-square test of independence |
| Does channel mix match specified shares? | Chi-square goodness-of-fit test |

A useful decision process is:

### Are You Comparing Two Population Proportions?

For example:

> Does the new page have a different conversion rate from the current page?

Use:

$$
\boxed{\text{Two-Proportion Inference}}
$$

---

### Are You Examining Two Categorical Variables?

For example:

> Is purchase outcome associated with which page a customer sees?

or:

> Is homeownership associated with marital status?

Use:

$$
\boxed{\text{Chi-Square Test of Independence}}
$$

---

### Are You Comparing One Categorical Variable to Predetermined Proportions?

For example:

> Does the observed channel mix match the expected 50/30/20 split?

Use:

$$
\boxed{\text{Chi-Square Goodness-of-Fit Test}}
$$

---

# Categorical Inference Workflow

When working with categorical data:

## Step 1: Identify the Business Question

Determine what comparison is actually being requested.

---

## Step 2: Identify the Observational Unit

Ask:

> What does one observation represent?

This is essential for determining whether observations are independent.

---

## Step 3: Identify the Categorical Variables

Determine whether the problem contains:

- One categorical variable
- Two categorical variables
- Two groups with a binary outcome

---

## Step 4: Examine Counts and Proportions

Construct:

- Frequency tables
- Contingency tables
- Conditional proportions
- Bar charts when useful

Do not interpret raw counts without considering group sizes.

---

## Step 5: Choose the Appropriate Method

$$
\begin{array}{c}
\text{Categorical Outcome}\\
\downarrow\\
\begin{cases}
\text{Two proportions}
\rightarrow
\text{Two-proportion inference}\\[6pt]
\text{Two categorical variables}
\rightarrow
\chi^2\text{ independence}\\[6pt]
\text{One categorical variable vs. benchmark}
\rightarrow
\chi^2\text{ goodness-of-fit}
\end{cases}
\end{array}
$$

---

## Step 6: State the Null Model

For two proportions:

$$
H_0:p_1=p_2
$$

For independence:

$$
H_0:
\text{The categorical variables are independent}
$$

For goodness-of-fit:

$$
H_0:
\text{The category probabilities equal the specified benchmark}
$$

---

## Step 7: Check Conditions

Depending on the method, examine:

- Independence of observations
- Mutually exclusive categories
- Expected counts
- Sample sizes

---

## Step 8: Conduct the Analysis

Use the appropriate R function:

```{r}
#| eval: false
prop.test()
```

or:

```{r}
#| eval: false
chisq.test()
```

---

## Step 9: Interpret in Context

Translate the statistical result back into the business question.

Avoid saying only:

> The p-value is significant.

Instead explain **what statistical relationship the data support**.

---

## Step 10: Separate Association from Causation

A statistically significant association does not automatically establish causality.

Ask:

> **How were the data generated?**

Random assignment can support causal interpretation.

Observational association by itself cannot.

---

# Connecting Lectures 5–8

The inferential methods covered so far can now be organized according to the **type of outcome and study design**.

| Question | Parameter / Structure | Method |
|---|---|---|
| What is one population mean? | $\mu$ | One-sample t |
| What is the mean difference for linked observations? | $\mu_d$ | Paired t |
| How do two independent population means differ? | $\mu_1-\mu_2$ | Welch two-sample t |
| How do two population proportions differ? | $p_1-p_2$ | Two-proportion inference |
| Are two categorical variables associated? | Contingency table | Chi-square independence |
| Does one categorical distribution match specified shares? | $p_1,\ldots,p_k$ | Chi-square goodness-of-fit |

The broader decision process is:

$$
\boxed{
\text{Business Question}
\rightarrow
\text{Outcome Type}
\rightarrow
\text{Study Design}
\rightarrow
\text{Parameter}
\rightarrow
\text{Statistical Method}
\rightarrow
\text{Interpretation}
}
$$

The key idea remains the same:

> **Do not start by choosing a statistical test. Start by understanding what the variables represent, how the data were generated, and what population question you are trying to answer.**