# Descriptive Analytics for Time Series Data

```{r}
#| label: setup
#| include: false
library(lubridate)
library(tidyverse)
library(ggplot2)
```

## Overview

For descriptive analysis, the focus was on questions such as: What is typical? How spread out is the data? What shape does the distribution have? With time series data, an additional question becomes important: When did it happen? 

Time series data is a sequence of observations collected or recorded at successive points in time. Because observations are ordered chronologically, time series data provides a dynamic view of how a variable changes over time, rather than a static snapshot.

- Chronological order: Observations occur sequentially over time so ordering of the data matters.
- Time index: Each oberservation is associated with a specific point in time.
- Dependence over time: Earlier and later observations may be related, meaning observations may not be independent.
- Temporal patterns: The ordering of observations allows patterns like trend, seasonality, and short term variation to be identified.

Suppose a CFO says"
    "Sales dropped 15% from Q3 to Q4."

That number by itself doesn't necessarily mean the business suddently got worse. Maybe sales normally decline from Q3 to Q4 and this is just a regular reoccurring observation.

With time series data, there are three things to be distinguished:

- Trend $\rightarrow$ Is there a long-term upward/downward movement?
- Seasonality $\rightarrow$ Does a pattern repeat at regular intervals?
- Noise $\rightarrow$ Is a movement just short-term random variation?
- Level Change $\rightarrow$ Did the series actually shift to a different level?

## Working with Lubridate

Why do dates matter in analytics? Dates can be used for several things, for example:

- Time aware filtering (e.g. time based filtering), time aware grouping (group a time period), and trend/seasonality detection.
- Aligning events (comparing before and afters).
- Preparing for forecasting.

Dates need to actually be dates! A date stored as text like: 

`'2025-08-21'`

is much different from an `R Date` object actually representing August 21, 2025. The main package used when working with dates is `lubridate`.

### Parsing Dates

Parsing dates is the process of converting raw string dates to an actual date/time object, which can then be converted into useful time components.
Raw string dates may not always be written in a consistent way, therefore different functions are used based on what the text format is.

- `ymd()` when string = YEAR $\rightarrow$ MONTH $\rightarrow$ DAY
- `mdy()` when string = MONTH $\rightarrow$ DAY $\rightarrow$ YEAR
- `dmy()` when string = DAY $\rightarrow$ MONTH $\rightarrow$ YEAR
- `ymd_hms()` when string = YEAR $\rightarrow$ MONTH $\rightarrow$ DAY `:` HH:MM:SS

Example:
```{r}
ymd("2025-08-21")
mdy("08/21/2025")
dmy("21-Aug-2025")
ymd_hms("2025-08-21 14:30:00")
```

Once the date is in the correct format, R can extract pieces:

- `year(x)` = extract year
- `month(x)` = extract month
- `quarter(x)` = extract quarter
- `wday(x)` = extract weekday
- `hour(x)` = extract hour
- `isoweek(x)` = extract week number of year

```{r}
x <- ymd_hms("2025-08-21 14:30:00")
x
year(x)
month(x)
quarter(x)
wday(x)
hour(x)
isoweek(x)

```

Esssentially the process is **raw date string → actual date/time object → useful time components**

Dates can also be created from combining variables that already exist. For example if you have `year = 2025`
and `month = 8` you can construct an actual date.

`make_date()` use when year/month columns exist because it avoids rounding problems associated with fractional-year representations

```{r}
# Before: year and month are separate variables
txhousing |>
    select(year, month) |>
    head(2)

# Create an actual date from year and month
TX <- txhousing |>
    mutate(
        date_actual = make_date(
            year = year,
            month = month,
            day = 1
        )
    )

# After: year and month have been combined into a date
TX |>
    select(year, month, date_actual) |>
    head(2)
```

### Calendar Arithmetic

Leap years can screw up calculation of dates. For example `ymd("2024-02-29")` is a leap year. If you subtract five years you'd get `2019-02-29` which doesn't exist. 

```{r}
d1 <- ymd("2024-02-29")
```

lubridate has calendar awareness operators:
```{r}
d1 %m-% years(5)
ymd("2023-01-31") %m+% months(1)
```

### Time Windows

Once dates are represented properly, normal comparisons work:

```{r}
filter(
    date_actual >= as.Date("2005-01-01"),
    date_actual <= as.Date("2010-12-31")
)
```

You can also make relative windows.

For example, find the most recent date:

```{r}
max_date <- max(txhousing$date_actual, na.rm = TRUE)

```

Then go five years backward:

```{r}
start_5y <- max_date %m-% years(5)
```

Then 

```{r}
recent <- txhousing |>
    filter(date_actual > start_5y)
```

So instead of hard-coding a five-year range, you’re saying:

Give me observations from the final five years of the dataset.

For time-series data, the standard visualization is a line chart.

The lecture calculates statewide median sales for each month:

```{r}
txhousing |>
    group_by(date_actual) |>
    summarise(
        median_sales = median(sales, na.rm = TRUE)
    ) |>
    ggplot(aes(date_actual, median_sales)) +
    geom_line()
```

The conceptual structure is:
Date → x-axis
Value → y-axis

Connecting observations makes the evolution of the variable visible.

`scale_x_date()` then controls how frequently dates are displayed so that a long time horizon doesn’t produce an unreadable x-axis.


### Seasonality

**Seasonality** is one of the central concepts of time series data.

Seasonality is a pattern that systematically repeats according to the calendar.

It can be viewed by:

Lines by year:
```{r}
ggplot(aes(
    x = month(date_actual, label = TRUE),
    y = sales,
    group = year,
    color = as.factor(year)
)) +
    geom_line()
```

Boxplots by month

What does a typical January look like across years? What does a typical July look like?

```{r}
ggplot(aes(month_lab, sales)) +
    geom_boxplot()
```

Heatmaps are another view of seasonality

       2005  2006  2007  2008 ...
Jan      ■     ■     ■     ■
Feb      ■     ■     ■     ■
Mar      ■     ■     ■     ■
...
Dec      ■     ■     ■     ■

where the fill intensity represents sales.

### Moving Averages

Month-to-month fluctuations can make the underlying trend difficult to see. A rolling average smooths those fluctuations.

```{r}
rollmean(
    median,
    k = 12,
    fill = NA,
    align = "right"
)
```

k = 12 means a 12-observation window. Because the data are monthly, that’s a 12-month moving average.
