# One-Parameter Statistical Inference

## Overview

So far, the focus has been on describing observed data and modeling uncertainty.

Descriptive statistics summarize the data we actually observe. Probability distributions describe how uncertain outcomes behave. **Statistical inference** takes the next step:

> **What can a sample tell us about a population we cannot fully observe?**

Suppose a company reports:

> *Average household spending last month was $1,793.*

That number may sound precise, but suppose it came from a sample of 2,500 households rather than every household.

The sample mean:

$$
\bar{x} = 1793
$$

is known, but the population mean:

$$
\mu
$$

is unknown.

The goal of statistical inference is to use the observed **sample statistic** to learn about the unknown **population parameter**, while explicitly accounting for sampling uncertainty.

The general process is:

$$
\text{Population}
\rightarrow
\text{Sample}
\rightarrow
\text{Statistic}
\rightarrow
\text{Inference about Parameter}
$$

There are two major tools for doing this:

- **Confidence intervals** → estimate a population parameter
- **Hypothesis tests** → evaluate a claim about a population parameter

Before either one works, however, we need to understand **sampling variability**.

---

# Sampling

## Why Sampling Matters

Usually, observing an entire population is impractical or impossible.

Instead, we collect a **sample** and use it to learn about the population.

For example:

- Population → all customers
- Sample → 2,500 selected customers
- Parameter → true average spending, $\mu$
- Statistic → sample average spending, $\bar{x}$

But the quality of an inference depends on **how the sample was obtained**.

A huge sample does not automatically mean a good sample.

Suppose a company surveys 50,000 customers about satisfaction, but only sends the survey to customers enrolled in its loyalty program.

Even though:

$$
n = 50,000
$$

the sample may systematically overrepresent highly engaged customers.

Increasing the sample size does not fix a poorly designed sampling process.

---

## Bias vs. Variance

Two major sources of uncertainty in sampling are **bias** and **variance**.

### Bias

**Bias** is a systematic tendency for a sampling process to produce estimates that differ from the true population value.

Bias generally comes from the **design of the data collection process**.

Examples include:

- Coverage bias
- Selection bias
- Self-selection bias
- Nonresponse bias

### Coverage Bias

**Coverage bias** occurs when the sampling frame fails to include some parts of the population.

For example, surveying only loyalty-program members excludes customers who never joined the program.

### Selection Bias

**Selection bias** occurs when the method used to select observations systematically favors certain types of observations.

Convenience samples are a common example.

### Self-Selection Bias

**Self-selection bias** occurs when individuals choose whether to participate.

For example, online reviews may disproportionately represent customers with extremely positive or extremely negative experiences.

### Nonresponse Bias

**Nonresponse bias** occurs when individuals who do not respond differ systematically from those who do.

For example, highly dissatisfied customers may be less willing to complete a company survey.

---

### Variance

**Variance** in sampling refers to the natural sample-to-sample fluctuation that occurs even when the sampling design is good.

Imagine randomly sampling 100 customers and calculating their average spending.

Then repeat the process with another 100 customers.

The two samples probably will not produce exactly the same mean.

That difference is **sampling variability**.

Increasing sample size reduces this variability.

The key distinction is:

| Problem | Meaning | Main Solution |
|---|---|---|
| **Bias** | Systematically wrong | Better sampling design |
| **Variance** | Naturally unstable across samples | Larger sample size |

A larger sample can reduce variance.

A larger sample **cannot fix systematic bias**.

> **Bad design + huge sample = precise estimate of the wrong thing.**

---

## Sampling Methods

### Simple Random Sampling

In **simple random sampling**, every member of the population has an equal chance of being selected.

Example:

> Randomly select transactions from a database for an audit.

Simple random sampling is conceptually straightforward and helps reduce selection bias.

---

### Stratified Sampling

In **stratified sampling**, the population is divided into meaningful subgroups called **strata**, and observations are sampled within each subgroup.

Examples of strata might include:

- Region
- Customer segment
- Age group
- Store type

For example, a company might sample customers separately from each geographic region so that all regions are appropriately represented.

---

### Cluster Sampling

In **cluster sampling**, entire groups are randomly selected rather than individual observations.

For example:

> Randomly select 20 stores and analyze customers from those stores.

Clusters might include:

- Stores
- Branches
- Schools
- Geographic areas

Cluster sampling can be cheaper and easier to implement, but observations within the same cluster may be similar, which can reduce precision.

---

### Systematic Sampling

In **systematic sampling**, observations are selected at regular intervals after a random starting point.

For example:

> Inspect every 20th item produced on an assembly line.

If the random starting point is item 7, the sample might contain:

$$
7,27,47,67,\ldots
$$

---

## Sampling Method Comparison

| Method | Main Idea | Example |
|---|---|---|
| Simple Random | Randomly select individuals | Randomly audit transactions |
| Stratified | Sample within known subgroups | Sample customers within each region |
| Cluster | Randomly select entire groups | Select stores, then observe customers |
| Systematic | Select every $k$th observation | Inspect every 20th product |

---

# Sampling Distributions

Suppose the population mean household spending is:

$$
\mu
$$

We randomly select a sample and calculate:

$$
\bar{x}
$$

If we selected another random sample, we would probably get a slightly different value of $\bar{x}$.

Suppose we repeated this process 1,000 times:

1. Draw a random sample of size $n$
2. Calculate $\bar{x}$
3. Store the result
4. Repeat 1,000 times

We would now have 1,000 different sample means.

The distribution of those sample means is called the **sampling distribution of the sample mean**.

### Sampling Distribution

A **sampling distribution** is the probability distribution of a statistic across repeated samples of the same size.

This is different from the distribution of the raw observations.

For example:

- Distribution of household spending → distribution of individual household values
- Sampling distribution of $\bar{x}$ → distribution of sample averages

That distinction is fundamental.

> **Population distribution:** How do individual observations vary?
>
> **Sampling distribution:** How would a statistic vary if we repeatedly sampled from the population?

---

## Center of the Sampling Distribution

For independent and identically distributed observations:

$$
X_1,X_2,\ldots,X_n
$$

with population mean:

$$
\mu
$$

the expected value of the sample mean is:

$$
E[\bar{X}] = \mu
$$

This means that the sampling distribution of $\bar{X}$ is centered at the true population mean.

Therefore, $\bar{X}$ is an **unbiased estimator** of $\mu$.

Across many repeated random samples, the sample mean does not systematically overestimate or underestimate the population mean.

---

# Standard Error

The spread of a sampling distribution is measured by the **standard error**.

For the sample mean:

$$
SE(\bar{X}) = \frac{\sigma}{\sqrt{n}}
$$

where:

$\sigma$ = population standard deviation  
$n$ = sample size

Usually, $\sigma$ is unknown, so we estimate it using the sample standard deviation:

$$
SE(\bar{X}) \approx \frac{s}{\sqrt{n}}
$$

The standard error therefore measures:

> **How much would the sample mean naturally vary from sample to sample?**

This is different from standard deviation.

### Standard Deviation vs. Standard Error

**Standard deviation** describes variation among **individual observations**:

$$
s
$$

**Standard error** describes variation in a **sample statistic across repeated samples**:

$$
SE(\bar{X})=\frac{s}{\sqrt{n}}
$$

| Measure | What Varies? | Interpretation |
|---|---|---|
| Standard Deviation | Individual observations | How spread out are the data? |
| Standard Error | Sample statistic | How precise is the estimate? |

---

## Effect of Sample Size on Standard Error

Because:

$$
SE(\bar{X})=\frac{s}{\sqrt{n}}
$$

increasing $n$ decreases the standard error.

Therefore:

$$
n \uparrow
\quad\Rightarrow\quad
SE \downarrow
\quad\Rightarrow\quad
\text{more precise estimate}
$$

However, the relationship involves the **square root** of $n$.

To cut the standard error in half, the sample size must be multiplied by four.

For example:

$$
SE=\frac{s}{\sqrt{100}}=\frac{s}{10}
$$

while:

$$
SE=\frac{s}{\sqrt{400}}=\frac{s}{20}
$$

So increasing the sample from 100 to 400 cuts the standard error in half.

---

# Central Limit Theorem

The **Central Limit Theorem (CLT)** explains why the Normal distribution becomes so important in statistical inference.

Suppose the original population is not Normal.

It could be:

- Right-skewed
- Left-skewed
- Bounded
- Irregular

As the sample size becomes sufficiently large, the **sampling distribution of the sample mean** becomes approximately Normal.

Mathematically:

$$
\bar{X}
\approx
N\left(
\mu,
\frac{\sigma^2}{n}
\right)
$$

for sufficiently large $n$.

Equivalently:

$$
E[\bar{X}]=\mu
$$

and:

$$
SE(\bar{X})=\frac{\sigma}{\sqrt{n}}
$$

The important idea is:

> **The raw data do not have to be Normally distributed for the sample mean to have an approximately Normal sampling distribution.**

---

## What the CLT Does Not Say

The Central Limit Theorem does **not** say that increasing the sample size makes the raw data Normal.

Suppose customer spending is strongly right-skewed.

Collecting more customers does not necessarily make the individual spending values bell-shaped.

Instead:

$$
\text{Raw data may remain skewed}
$$

while:

$$
\text{Distribution of sample means becomes approximately Normal}
$$

This distinction is critical.

The lecture demonstrates this using a badly right-skewed population. Sample means with $n=3$ remain visibly skewed, while the sampling distributions at $n=30$ and especially $n=100$ become increasingly bell-shaped.

---

## How Large Does $n$ Need to Be?

A common rule of thumb is:

$$
n \geq 30
$$

However, this is only a guideline.

The amount of data needed depends on the underlying population.

A population with:

- Strong skewness
- Heavy tails
- Extreme outliers

may require a substantially larger sample before the sampling distribution becomes approximately Normal.

---

## Why the CLT Matters

The Central Limit Theorem allows us to use Normal-based statistical methods even when the underlying population itself is not Normal.

This provides the foundation for:

- Confidence intervals
- Hypothesis tests
- A/B testing
- Estimating business KPIs
- Operational thresholds

The conceptual chain is:

$$
\text{Random Sample}
\rightarrow
\text{Sampling Distribution}
\rightarrow
\text{CLT}
\rightarrow
\text{Standard Error}
\rightarrow
\text{Statistical Inference}
$$

---

# Inference for a Population Mean

Suppose we want to estimate an unknown population mean:

$$
\mu
$$

We observe a sample and calculate:

$$
\bar{x}
$$

The sample mean is our **point estimate** of the population mean.

$$
\boxed{\text{Point Estimate}=\bar{x}}
$$

But a point estimate alone does not communicate uncertainty.

For example:

> Average household spending is estimated to be $1,793.

How precise is that estimate?

Could the true population mean reasonably be $1,790?

What about $1,500?

To quantify this uncertainty, we construct a **confidence interval**.

---

# Confidence Intervals

A **confidence interval** gives a range of plausible values for an unknown population parameter.

The general structure is:

$$
\boxed{
\text{Estimate}
\pm
\text{Margin of Error}
}
$$

and:

$$
\text{Margin of Error}
=
\text{Critical Value}
\times
\text{Standard Error}
$$

Therefore:

$$
\boxed{
\text{Confidence Interval}
=
\text{Estimate}
\pm
(\text{Critical Value})(\text{SE})
}
$$

---

## Confidence Interval for a Mean

When estimating $\mu$, the population standard deviation $\sigma$ is usually unknown.

Therefore, we substitute:

$$
s
$$

for:

$$
\sigma
$$

and use the **t distribution** rather than the standard Normal distribution.

The confidence interval is:

$$
\boxed{
\bar{x}
\pm
t_{\alpha/2,n-1}
\frac{s}{\sqrt{n}}
}
$$

where:

$\bar{x}$ = sample mean  
$s$ = sample standard deviation  
$n$ = sample size  
$n-1$ = degrees of freedom  
$t_{\alpha/2,n-1}$ = critical value from the t distribution

For a 95% confidence interval:

$$
\alpha=0.05
$$

so the critical value corresponds to:

$$
1-\frac{\alpha}{2}
=
0.975
$$

---

# The t Distribution

The **t distribution** is similar to the standard Normal distribution but has **heavier tails**.

Why?

Because when $\sigma$ is unknown, we estimate it using:

$$
s
$$

That introduces additional uncertainty.

Therefore:

$$
\frac{\bar{X}-\mu}{s/\sqrt{n}}
\sim
t_{n-1}
$$

rather than:

$$
N(0,1)
$$

The degrees of freedom are:

$$
df=n-1
$$

As the sample size increases:

$$
t_{n-1}
\rightarrow
N(0,1)
$$

So for large samples, the t and z critical values become nearly identical.

---

## Confidence Level

For a 95% confidence interval:

$$
1-\alpha=0.95
$$

so:

$$
\alpha=0.05
$$

Because a two-sided interval splits this probability across both tails:

$$
\frac{\alpha}{2}=0.025
$$

Therefore the middle 95% of the reference distribution lies between the two critical values.

---

## Interpreting a Confidence Interval

Suppose the estimated 95% confidence interval for average household spending is:

$$
[1716,1869]
$$

A standard interpretation is:

> **We are 95% confident that the interval from $1,716 to $1,869 contains the population mean household spending.**

The precise frequentist interpretation refers to the **procedure**, not probability assigned to the fixed parameter.

If we repeatedly:

1. Drew random samples
2. Constructed a 95% confidence interval from each sample

then approximately 95% of those intervals would contain the true population mean.

Therefore, technically, we should not say:

> There is a 95% probability that $\mu$ is inside this specific interval.

Once the interval has been calculated, both its endpoints and the population parameter are fixed. The 95% refers to the **long-run success rate of the procedure**.

---

## Confidence Interval Example: Household Spending

Suppose the sample produces:

$$
\bar{x}=1792.77
$$

$$
s=1947.92
$$

and:

$$
SE=39.12
$$

The resulting 95% confidence interval is approximately:

$$
[1716,1869]
$$

So instead of reporting only:

> Average household spending is $1,793.

we can report:

> Average household spending is estimated at approximately $1,793, with a 95% confidence interval from approximately $1,716 to $1,869.

The second statement communicates both the estimate and its uncertainty.

---

## Confidence Interval for a Mean in R

R can calculate the confidence interval directly using:

```{r}
t.test(x)
```

The calculation can also be performed manually:

```{r}
mean(x) +
    c(-1, 1) *
    qt(0.975, length(x) - 1) *
    sd(x) / sqrt(length(x))
```

The structure corresponds directly to:

$$
\bar{x}
\pm
t^*
\frac{s}{\sqrt{n}}
$$

---

## What Determines Confidence Interval Width?

The width of a confidence interval depends primarily on:

- Sample size
- Variability
- Confidence level

### Larger Sample Size

$$
n\uparrow
\Rightarrow
SE\downarrow
\Rightarrow
\text{narrower CI}
$$

### Greater Variability

$$
s\uparrow
\Rightarrow
SE\uparrow
\Rightarrow
\text{wider CI}
$$

### Higher Confidence Level

Increasing the confidence level requires a larger critical value.

Therefore:

$$
\text{Confidence Level}\uparrow
\Rightarrow
\text{Critical Value}\uparrow
\Rightarrow
\text{wider CI}
$$

There is therefore a tradeoff between **confidence** and **precision**.

---

# Hypothesis Testing

A **hypothesis test** evaluates whether sample evidence is sufficiently inconsistent with a claim about a population parameter.

Instead of asking:

> What values of $\mu$ are plausible?

as in a confidence interval, a hypothesis test asks:

> Is the sample evidence strong enough to challenge a specific claim about $\mu$?

---

## Null and Alternative Hypotheses

A hypothesis test begins with two competing hypotheses.

### Null Hypothesis

The **null hypothesis**, $H_0$, represents the baseline or status-quo claim.

For example:

$$
H_0:\mu=325
$$

### Alternative Hypothesis

The **alternative hypothesis**, $H_a$, represents the claim being investigated.

For an increase:

$$
H_a:\mu>325
$$

For a decrease:

$$
H_a:\mu<325
$$

For any difference:

$$
H_a:\mu\neq325
$$

The form of the alternative determines whether the test is:

| Alternative | Test |
|---|---|
| $H_a:\mu>\mu_0$ | Right-tailed |
| $H_a:\mu<\mu_0$ | Left-tailed |
| $H_a:\mu\neq\mu_0$ | Two-tailed |

---

# Hypothesis Testing Workflow

A hypothesis test follows six major steps:

1. State the hypotheses
2. Choose the significance level, $\alpha$
3. Calculate the test statistic
4. Calculate the p-value
5. Make a statistical decision
6. State the conclusion in context

The process can be summarized as:

$$
H_0/H_a
\rightarrow
\alpha
\rightarrow
\text{Test Statistic}
\rightarrow
p\text{-value}
\rightarrow
\text{Decision}
\rightarrow
\text{Conclusion}
$$

---

# Test Statistic for a Mean

For a one-sample test of:

$$
H_0:\mu=\mu_0
$$

the test statistic is:

$$
\boxed{
t=
\frac{\bar{x}-\mu_0}
{s/\sqrt{n}}
}
$$

The numerator:

$$
\bar{x}-\mu_0
$$

measures the observed difference between the sample mean and the hypothesized population mean.

The denominator:

$$
\frac{s}{\sqrt{n}}
$$

measures the expected sampling variability.

Therefore, the t statistic can be interpreted as:

> **How many standard errors away is the observed sample mean from the value claimed by the null hypothesis?**

Under $H_0$:

$$
t\sim t_{n-1}
$$

---

# P-Values

The **p-value** measures how surprising the observed result would be if the null hypothesis were true.

More precisely:

> The p-value is the probability, assuming $H_0$ is true, of obtaining a test statistic at least as extreme as the one observed in the direction(s) specified by $H_a$.

A small p-value means the observed data would be relatively unusual under the null hypothesis.

The standard decision rule is:

$$
p \leq \alpha
\Rightarrow
\text{Reject }H_0
$$

and:

$$
p > \alpha
\Rightarrow
\text{Fail to reject }H_0
$$

---

## What a P-Value Is Not

A p-value is **not**:

$$
P(H_0\text{ is true})
$$

It is also not:

$$
P(H_a\text{ is true})
$$

The p-value is calculated **assuming the null hypothesis is true**.

It tells us how compatible the observed data are with that assumption.

---

# Reject vs. Fail to Reject

Suppose:

$$
\alpha=0.05
$$

If:

$$
p=0.02
$$

then:

$$
p<\alpha
$$

so we **reject $H_0$**.

If:

$$
p=0.13
$$

then:

$$
p>\alpha
$$

so we **fail to reject $H_0$**.

Importantly:

> **Fail to reject does not mean accept $H_0$.**

It means the available sample does not provide sufficient evidence against $H_0$ at the chosen significance level.

Absence of sufficient evidence is not proof that the null hypothesis is true.

---

# Worked Example: Food Expenditure

Suppose last year's average monthly food expenditure was:

$$
\$325
$$

A researcher wants to determine whether average spending has increased.

The sample contains:

$$
n=25
$$

$$
\bar{x}=355
$$

$$
s=128
$$

The hypotheses are:

$$
H_0:\mu=325
$$

$$
H_a:\mu>325
$$

The test statistic is:

$$
t=
\frac{355-325}
{128/\sqrt{25}}
$$

$$
t\approx1.172
$$

The resulting p-value is:

$$
p\approx0.126
$$

At:

$$
\alpha=0.05
$$

we have:

$$
0.126>0.05
$$

Therefore:

> **Fail to reject $H_0$.**

There is not sufficient statistical evidence at the 5% significance level to conclude that mean monthly food expenditure has increased above $325.

Notice that:

$$
\bar{x}=355>325
$$

but the result is still not statistically significant.

The observed difference must be considered relative to its **sampling uncertainty**.

---

## Hypothesis Test in R

The test can be performed manually:

```{r}
t_statistic <- (355 - 325) / (128 / sqrt(25))

p_value <- 1 - pt(
    t_statistic,
    df = 24
)

t_statistic
p_value
```

Or directly:

```{r}
t.test(
    FoodCosts$Monthly_Food_Cost,
    alternative = "greater",
    mu = 325
)
```

---

# Statistical Significance vs. Practical Significance

A statistically significant result does not automatically mean the effect is important in practice.

Suppose a hospital claims that average ER waiting time is:

$$
10\text{ minutes}
$$

A sample produces:

$$
n=40
$$

$$
\bar{x}=11
$$

$$
s=3
$$

Testing:

$$
H_0:\mu=10
$$

against:

$$
H_a:\mu>10
$$

produces approximately:

$$
t=2.11
$$

and:

$$
p=0.021
$$

Since:

$$
p<0.05
$$

the result is statistically significant.

But the estimated difference is only:

$$
11-10=1\text{ minute}
$$

Whether that one-minute difference is operationally important is a **business or practical question**, not something determined solely by the p-value.

Therefore:

> **Statistical significance asks whether an observed effect is difficult to explain by sampling variation under $H_0$.**
>
> **Practical significance asks whether the size of the effect actually matters.**

Both should be considered.

---

# Errors in Hypothesis Testing

Because hypothesis tests make decisions using uncertain samples, errors are possible.

There are two major types.

## Type I Error

A **Type I error** occurs when we reject the null hypothesis even though it is actually true.

$$
P(\text{Type I Error})=\alpha
$$

Conceptually:

> **False positive**

Example:

> Concluding a new marketing strategy increases sales when it actually does not.

---

## Type II Error

A **Type II error** occurs when we fail to reject the null hypothesis even though the alternative is actually true.

$$
P(\text{Type II Error})=\beta
$$

Conceptually:

> **False negative**

Example:

> Failing to detect that a marketing strategy really does increase sales.

---

## Decision Table

| Reality | Fail to Reject $H_0$ | Reject $H_0$ |
|---|---|---|
| $H_0$ true | Correct decision | **Type I Error** |
| $H_a$ true | **Type II Error** | Correct detection |

---

# Significance Level

The significance level:

$$
\alpha
$$

controls the probability of a Type I error.

A common choice is:

$$
\alpha=0.05
$$

meaning the testing procedure allows a 5% probability of rejecting $H_0$ when it is actually true.

Reducing $\alpha$ makes the test more conservative.

However, for a fixed sample size, reducing $\alpha$ generally increases the chance of a Type II error.

Therefore, there is a tradeoff:

$$
\alpha\downarrow
\Rightarrow
\beta\uparrow
$$

when the other components of the test remain fixed.

---

# Statistical Power

**Power** is the probability that a hypothesis test correctly rejects the null hypothesis when the alternative hypothesis is true.

$$
\boxed{
\text{Power}
=
P(\text{Reject }H_0\mid H_a\text{ true})
}
$$

Since:

$$
\beta
=
P(\text{Fail to reject }H_0\mid H_a\text{ true})
$$

we have:

$$
\boxed{
\text{Power}=1-\beta
}
$$

Power measures the ability of a statistical test to **detect a real effect**.

---

## What Determines Power?

Power depends on:

- Sample size
- Effect size
- Variability
- Significance level

### Larger Sample Size

$$
n\uparrow
\Rightarrow
SE\downarrow
\Rightarrow
\text{Power}\uparrow
$$

### Larger Effect

A larger true difference from the null value is easier to detect.

$$
|\mu_1-\mu_0|\uparrow
\Rightarrow
\text{Power}\uparrow
$$

### Lower Variability

Less noisy data make real differences easier to distinguish.

$$
\sigma\downarrow
\Rightarrow
\text{Power}\uparrow
$$

### Significance Level

For fixed $n$:

$$
\alpha\downarrow
\Rightarrow
\text{harder to reject }H_0
\Rightarrow
\text{Power}\downarrow
$$

---

## Food Expenditure Power Example

In the earlier example:

$$
n=25
$$

and:

$$
s=128
$$

Suppose the true mean were actually:

$$
\mu_1=350
$$

instead of the null value:

$$
\mu_0=325
$$

The lecture's power calculation gives approximately:

$$
\text{Power}=0.243
$$

or about:

$$
24\%
$$

So even if the true population mean really were $350, a study of this size and variability would detect the increase only about 24% of the time using this testing setup.

That helps explain why:

> **Failing to reject $H_0$ does not necessarily mean there is no real effect.**

The study may simply have low power.

---

# Inference for a Population Proportion

The same inferential framework can be applied when the variable is **binary** rather than quantitative.

Suppose each customer either:

$$
X_i=
\begin{cases}
1 & \text{customer converts}\\
0 & \text{customer does not convert}
\end{cases}
$$

Then:

$$
X_i\sim\text{Bernoulli}(p)
$$

where:

$$
p
$$

is the true population probability of conversion.

The sample proportion is:

$$
\hat{p}
=
\frac{1}{n}
\sum_{i=1}^{n}X_i
$$

Because the values are 0 and 1, the sample proportion is literally the **sample mean of binary observations**.

---

# Sampling Distribution of a Proportion

For a population proportion $p$:

$$
E[\hat{p}]=p
$$

and:

$$
SE(\hat{p})
=
\sqrt{
\frac{p(1-p)}{n}
}
$$

For sufficiently large samples, the Central Limit Theorem gives:

$$
\frac{
\hat{p}-p
}{
\sqrt{p(1-p)/n}
}
\approx
N(0,1)
$$

Therefore, the sampling distribution of $\hat{p}$ is approximately Normal when the sample is sufficiently large.

---

# Confidence Interval for a Proportion

Because the true population proportion $p$ is unknown, we estimate the standard error using:

$$
\hat{p}
$$

Therefore:

$$
SE(\hat{p})
\approx
\sqrt{
\frac{
\hat{p}(1-\hat{p})
}{
n
}
}
$$

A confidence interval is:

$$
\boxed{
\hat{p}
\pm
z_{\alpha/2}
\sqrt{
\frac{
\hat{p}(1-\hat{p})
}{
n
}
}
}
$$

For a 95% confidence interval:

$$
z^*\approx1.96
$$

---

## Example: E-Commerce Conversion Rate

Suppose an e-commerce company observes whether visitors convert:

- `1` = converted
- `0` = did not convert

Then:

```{r}
x <- subset(
    ab_data,
    landing_page == "old_page"
)$converted
```

The sample conversion rate is:

```{r}
p_hat <- mean(x)
```

The estimated standard error is:

```{r}
se_p_hat <- sqrt(
    p_hat * (1 - p_hat) / length(x)
)
```

The 95% confidence interval is:

```{r}
p_hat +
    c(-1, 1) *
    qnorm(0.975) *
    se_p_hat
```

Interpretation:

> **We are 95% confident that the resulting interval contains the true conversion rate for old-page visitors.**

Just as with the mean:

$$
n\uparrow
\Rightarrow
SE\downarrow
\Rightarrow
\text{narrower CI}
$$

---

# Hypothesis Test for a Proportion

Suppose we want to test:

$$
H_0:p=p_0
$$

against an alternative such as:

$$
H_a:p>p_0
$$

The test statistic is:

$$
\boxed{
z=
\frac{
\hat{p}-p_0
}{
\sqrt{
p_0(1-p_0)/n
}
}
}
$$

Notice an important difference.

For the **confidence interval**, the standard error uses:

$$
\hat{p}
$$

because we are estimating the unknown population proportion.

For the **hypothesis test**, the standard error uses:

$$
p_0
$$

because the calculation asks what sampling variability would look like **assuming the null hypothesis is true**.

Therefore:

### Confidence Interval

$$
SE=
\sqrt{
\frac{
\hat{p}(1-\hat{p})
}{
n
}
}
$$

### Hypothesis Test

$$
SE_0=
\sqrt{
\frac{
p_0(1-p_0)
}{
n
}
}
$$

That distinction is important.

---

# Why Proportions Use z Instead of t

Inference for a mean uses the t distribution because the population standard deviation:

$$
\sigma
$$

is unknown and must be estimated using:

$$
s
$$

For a Bernoulli variable:

$$
Var(X)=p(1-p)
$$

so the variance is determined directly by the proportion itself.

Therefore, proportion inference uses the **Normal/z reference distribution** rather than a t distribution.

| Parameter | Reference Distribution |
|---|---|
| Mean $\mu$ | t |
| Proportion $p$ | z / Normal |

---

# Approximate vs. Exact Proportion Inference

The Normal-based confidence interval for a proportion relies on a **large-sample approximation**.

That approximation may become unreliable when:

- $n$ is small
- $\hat{p}$ is very close to 0
- $\hat{p}$ is very close to 1
- There are very few successes or failures

In those situations, an exact Binomial method may be preferable.

In R:

```{r}
binom.test(
    sum(ab_data$converted),
    length(ab_data$converted),
    conf.level = 0.95
)
```

`binom.test()` performs inference using the Binomial distribution rather than relying entirely on the Normal approximation.

---

# Mean vs. Proportion Inference

The formulas differ, but the underlying logic is the same.

| | Population Mean $\mu$ | Population Proportion $p$ |
|---|---|---|
| **Point Estimate** | $\bar{x}$ | $\hat{p}$ |
| **Estimated SE for CI** | $\frac{s}{\sqrt{n}}$ | $\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$ |
| **Reference Distribution** | $t_{n-1}$ | $z$ / Normal |
| **Confidence Interval** | $\bar{x}\pm t^*SE$ | $\hat{p}\pm z^*SE$ |
| **Null Hypothesis** | $H_0:\mu=\mu_0$ | $H_0:p=p_0$ |
| **Test Statistic** | $\frac{\bar{x}-\mu_0}{s/\sqrt{n}}$ | $\frac{\hat{p}-p_0}{\sqrt{p_0(1-p_0)/n}}$ |

Despite the different formulas, both follow the same basic process:

$$
\boxed{
\text{Estimate}
\rightarrow
\text{Quantify Uncertainty}
\rightarrow
\text{Compare to Reference Distribution}
\rightarrow
\text{Draw Conclusion}
}
$$

---

# Confidence Intervals vs. Hypothesis Tests

Confidence intervals and hypothesis tests are two ways of performing statistical inference.

### Confidence Interval

A confidence interval asks:

> **What values of the population parameter are reasonably compatible with the sample?**

The result is a range:

$$
[\text{Lower Bound},\text{Upper Bound}]
$$

### Hypothesis Test

A hypothesis test asks:

> **Is the sample sufficiently inconsistent with a specific hypothesized population value?**

The result involves:

- Test statistic
- p-value
- Reject or fail to reject decision
- Contextual conclusion

Both rely on the same underlying concepts:

- Sampling distributions
- Standard errors
- Reference distributions
- Sampling uncertainty

---

# Overall Inference Framework

The central idea of this lecture is that **sample statistics vary**.

A sample statistic such as:

$$
\bar{x}
$$

or:

$$
\hat{p}
$$

is not expected to equal the population parameter exactly.

Instead:

$$
\text{Observed Statistic}
=
\text{Population Signal}
+
\text{Sampling Noise}
$$

Statistical inference provides a framework for determining how much uncertainty that sampling noise creates.

The full workflow is:

$$
\text{Population Question}
\rightarrow
\text{Sampling Design}
\rightarrow
\text{Sample}
\rightarrow
\text{Point Estimate}
\rightarrow
\text{Standard Error}
\rightarrow
\text{Reference Distribution}
\rightarrow
\begin{cases}
\text{Confidence Interval}\\
\text{Hypothesis Test}
\end{cases}
\rightarrow
\text{Business Conclusion}
$$

The major ideas connect as follows:

- **Sampling design** determines whether the sample is trustworthy.
- **Bias** creates systematic error and cannot be fixed simply by increasing $n$.
- **Sampling variability** causes statistics to differ naturally across samples.
- **Standard error** quantifies that sample-to-sample variability.
- **The Central Limit Theorem** explains why many sampling distributions become approximately Normal.
- **Confidence intervals** use the estimate and its standard error to quantify plausible values for the population parameter.
- **Hypothesis tests** compare the observed estimate with a hypothesized parameter value.
- **P-values** quantify how unusual the observed evidence would be under $H_0$.
- **Type I and Type II errors** describe the two possible inferential mistakes.
- **Power** measures the probability of detecting an effect when one actually exists.
- The same overall logic applies to both **means** and **proportions**.